---
title: "Life Insurance"
output: html_document
date: "2023-01-22"
---

```{r install-packages, include=FALSE}
# Install packages only if not already installed
packages <- c("rlang", "caret", "dplyr", "ggplot2", "readr")
installed_packages <- rownames(installed.packages())
for (pkg in packages) {
  if (!pkg %in% installed_packages) {
    install.packages(pkg)
  }
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Cash Flows

## Scenario 1
Suppose you want to go on a trip 5 years from now. You have expenses of 3000 each year for 4 years (including this year), followed by an expense of 4000 and a final expense of 5000. How much would you have to deposit at this moment in time to cover all of
these future payments? Assume a constant interest rate of 2.14% and payments would take place at the beginning of the year.

```{r}
# We have an expense of 3000 from t=0 to t=3, an expense of 4000 at t=4 and 5000 at t=5

# Cash flows
cash_flows <- c(rep(3000,4), 4000, 5000)
                
# Interest Rate
i <- 0.0214

# Discount factor
v <- 1 / (1+i)

# Discount factors
discount_factors <- v ^ (0:5)

# Present value
present_val <- sum(cash_flows * discount_factors)
present_val
```
So you would have to make a deposit of 19800.98 at this current time to cover all of these future payments.

## Scenario 2
You're the head of a lawn mowing company. You want to invest in a new lawn mower than you are confident will generate cash
flows of 50 each month for 3 months, followed by 20 for 2 months. The issue is it costs 185 and you're unsure if it will be a profitable investment. To find out if it is a good investment or not, you must calculate the NPV for this investment. Assume the monthly interest rate is 0.6%.

```{r}
# Define the cash flows
# We have a negative cash flow at t=0 of 500, and 5 positive cash flows of 50 from t=1 to t=5

cash_flows <- c(0, rep(50,3), rep(30,2))

# Discount factors
discount_factors <- (1+0.006) ^ - (0:5)

# NPV
npv <- sum(cash_flows * discount_factors) - 185
npv

```
Since our NPV is positive, this indicates that this would be an overall gain ie profit.

## Scenario 3
Suppose instead of making those expenses, you chose to put them into a savings account. What would be your new present value, assuming same interest rate. What would be your new accumulated value?

```{r}

# Cash flows
cash_flows <- c(rep(3000,4), 4000, 5000)

# Define the discount function 
discount <- function(s, t, i = 0.0214) {
  (1 + i) ^ - (t - s)
}

# Calculate the pv
pv <- sum(cash_flows * discount(0, 0:5))
pv

# Calculate the value at time 6
av <- sum(cash_flows * discount(6, 0:5))

# Calculate the value at time 6, starting from present_value
pv * discount(6, 0)

print(av - pv)
```
As you can see, there would be a difference of $2682.41 if you saved the money instead of spending it.

## Scenario 4
Your parents want to start saving money for your university costs. Each year of study will set you guys back by about 7500. You're currently 14 (t=0) and will start attending when you are 19 (t=5). Therefore, your parents will make 1 deposit each year for 4 years (t=1 to t=4). Assume you will attend for 4 years (t=5 to t=22), the constant annual interest rate is 3.14% and each deposit your parents make is equal. Find the amount of the deposit

```{r}

# discount factors
discount_factors <- (1 + 0.0314) ^ - (0:8)

# deposit pattern
deposits <- c(0, rep(1, 4), rep(0, 4))

# university expenses
payments <- c(rep(0, 5), rep(7500, 4))

# Calculate the present value of the deposits
PV_deposit <- sum(deposits * discount_factors)

# Calculate the present value of the payments
PV_payment <- sum(payments * discount_factors)

# Calculate the yearly deposit K in the first 4 years
yearly_deposit <-  PV_payment / PV_deposit 
yearly_deposit

```
So your parents would have to make a deposit of 6627.546 for each of the 4 years to finance your university.

## Scenario 5
You want to take out a loan to open a new restaurant. You plan to take 1000 this year and 3000 next year. You plan to repay the money with equal yearly payments for the 10 years after next year (t=2 to t=11). Note this time the interest rate isn't constant but changes over time. For the first 3 years it is 4%, then for the next 3 years it is 4.5%, then for the final 5 years it is 6%. How much will your yearly payments be? 

```{r}

# Interest rates
interest <- c(rep(0.04, 3), rep(0.045, 3), rep(0.06, 5))

# Yearly discount factors
yearly_discount_factors <- (1 + interest) ^ (-1)

# Discount factors
discount_factors <- cumprod(c(1, yearly_discount_factors))

# Cash flows for the first two years
cash_flow <- c(1000, 3000)

# Calculate the present value (PV) of the loan
PV_loan <- sum(cash_flow * discount_factors[1:2])

# Calculate the present value (PV) of the repayments
repayment_years <- 2:11
discount_factors_repayments <- discount_factors[repayment_years + 1]
n_payments <- length(discount_factors_repayments)

# Define the yearly payment variable
yearly_payment <- PV_loan / sum(discount_factors_repayments)

# Create the full cash flow vector with the repayments
cash_flow_full <- c(cash_flow, rep(-yearly_payment, n_payments))

# Calculate the PV of the repayments to check the balance
PV_repayments <- sum(cash_flow_full * discount_factors)

# Display the yearly payment and the PV balance check
yearly_payment
PV_repayments
```
Our PV is ~0, meaning the PV of the future payments and PV of your loans are equal, hence the yearly payments would be 515.2757

## Scenario 6
You plan to take out a 100,000 loan with a yearly interest rate of 2.88%. You want to pay back this loan with fixed monthly payments over the next 15 years. Find how much these monthly payments will be.

```{r}

#number of payments
number_payments <- 12*15

#yearly interest rate
i <- 0.0288

#monthly interest rate

monthly_interest <- (1+i)^(1/12)-1

# Define the discount factors
discount_factors <- (1 + monthly_interest) ^ - (1:number_payments)

# Define the payment pattern
payments <- rep(1, number_payments)

# Calculate the monthly loan payment K
K <- 100000 / sum(payments * discount_factors)
K
```
These monthly payments will be 683.0401

# Life Tables

Import the life tables from mortality.org. We are going to import the dataset of Canada from 1921 - 2021.
```{r}
life_table <- read.table("C:\\Users\\megacrazyleo\\Desktop\\SQL\\R\\bltper_1x1.txt", header=FALSE, skip = 3, sep = "") 
colnames(life_table) <- c("Year", "Age", "mx", "qx", "ax", "lx", "dx", "Lx", "Tx", "ex")
```


Setting up variables 
```{r}
year <- life_table$Year
age <- life_table$Age
qx <- life_table$qx
mx <- life_table$mx
ax <- life_table$ax
lx <- life_table$lx
dx <- life_table$dx
Lx <- life_table$Lx
Tx <- life_table$Tx
ex <- life_table$ex
px <- 1 - qx
```
#### Year: 
Year or range of years (for both period & cohort data)

#### Age:
Age group for n-year interval from exact age x to just before exact age x+n, where n=1, 4, 5, or ∞ (open age interval)

#### m(x)
Central death rate between ages x and x+n

#### q(x)
Probability of death between ages x and x+n

#### a(x)
Average length of survival between ages x and x+n for persons dying in the interval

#### l(x)
Number of survivors at exact age x, assuming l(0) = 100,000

#### d(x)
Number of deaths between ages x and x+n

#### L(x)
Number of person-years lived between ages x and x+n

#### T(x)
Number of person-years remaining after exact age x

#### e(x)
Life expectancy at exact age x (in years)



## Mortality Rates
```{r}
# Setting up the plot
par(xpd = TRUE, mar = c(5, 4, 4, 6) + 0.1) # To ensure plot doesn't cover data

convert_age <- function(x) {
  x[x == "110+"] <- "110"  # Age is character in our dataset, need to change to numeric
  suppressWarnings({
    as.numeric(x)
  })
}
age <- convert_age(age)

# Create a color palette
years <- unique(year)
color_palette <- colorRampPalette(c("blue", "green", "yellow", "red"))(length(years))

# Initialize plot
plot2 <- plot(NULL, 
              xlim = range(age, na.rm = TRUE),  # Exclude NAs from range calculation
              ylim = range(qx, na.rm = TRUE),   # Exclude NAs from range calculation
              main = "Mortality rates (Canada, Males & Females, 1921-2021)",
              xlab = "Age x",
              ylab = expression(paste("Mortality rate ", q[x])),
              type = "l"
)

# Plot each year's data
suppressWarnings({
  for (i in seq_along(years)) {
    year_data <- subset(life_table, Year == years[i])
    lines(year_data$Age, year_data$qx, col = color_palette[i], type = "l")
  }
})

# Colored legend
legend("topright", inset = c(-0.2,0),
       legend = c("1921", "1955", "1988", "2021"),
       col = color_palette[c(1, which.min(abs(years - 1955)), 
                             which.min(abs(years - 1988)), length(years))],
       lty = 1, 
       title = "Year")

```

We can see that mortality rates from recent years are much lower compared to the earliest years. 

## Life Expectancy
```{r}

# Setting up the plot
par(xpd = TRUE, mar = c(5, 4, 4, 6) + 0.1) # To ensure plot doesn't cover data

# Create a color palette
years <- unique(year)
color_palette <- colorRampPalette(c("blue", "green", "yellow", "red"))(length(years))

plot3 <- suppressWarnings({
  plot(NULL, 
       xlim = range(age, na.rm = TRUE),  # Exclude NAs from range calculation
       ylim = range(ex, na.rm = TRUE),   # Exclude NAs from range calculation
       main = "Life Expectancy (Canada, Males & Females, 1921-2021)",
       xlab = "Age x",
       ylab = expression(paste("Life Expectancy ", e[x])),
       type = "l"
  )
})

# Plot each year's data
suppressWarnings({
  for (i in seq_along(years)) {
    year_data <- subset(life_table, Year == years[i])
    lines(year_data$Age, year_data$ex, col = color_palette[i], type = "l")
  }
})

# Colored legend
legend("topright", inset = c(-0.2, 0),
       legend = c("1921", "1955", "1988", "2021"),
       col = color_palette[c(1, which.min(abs(years - 1955)), 
                             which.min(abs(years - 1988)), length(years))],
       lty = 1, 
       title = "Year")



```

As we can see, the life expectancy is much higher from the years ~ 2021 as opposed to the years from ~1921 or even ~1955. 
Note for the early years, the life expectancy starts extremely low at age 0 and after a couple years, has a sharp increase. 
This can be explained mainly because it was much more difficult for an infant to survive the initial years. In later years, thanks
to modern medicine, the survival rate of infants have risen much higher.

Let us focus on the year 2021.

## 2021 Life Table
```{r}
life_table <- life_table[year == 2021, ]  #Getting only data from the year 2021
year <- life_table$Year
age <- life_table$Age
qx <- life_table$qx
mx <- life_table$mx
ax <- life_table$ax
lx <- life_table$lx
dx <- life_table$dx
Lx <- life_table$Lx
Tx <- life_table$Tx
ex <- life_table$ex
px <- 1 - qx
```

###Finding the probability that an 23 year old dies before turning 24
```{r}

qx[23 + 1]

```
As we can see, it is a very low probability, < 0.098%

### The expected future lifetime of an 23 year old
```{r}
ex[23 + 1]

```
It is expected that an average 23 year old from Canada will live for 57.06 more years in Canada in 2021


### Plot the mortality rates in the year 2021
```{r}
plot(age, qx, 
    main = "Mortality rates (Canada, 2021)", 
    xlab = "Age x", 
    ylab = expression(paste("Mortality rate ", q[x])), 
    type = "l")
```


### Plot the logarithm of the mortality rates in the year 2021
```{r}
# Plot the logarithm of the mortality rates in the year 2021
plot(age, log(qx), 
    main = "Log mortality rates (Canada, 2021)", 
    xlab = "Age x", 
    ylab = expression(paste("Log mortality rate ", log(q[x]))), 
    type = "l")
```

## Plot the mortality rates of females vs males in the year 2021

#### Import the male/female datasets
```{r}
#male data
life_table_m <- read.table("C:\\Users\\megacrazyleo\\Desktop\\SQL\\R\\mltper_1x1.txt", header=FALSE, skip = 3, sep = "")
colnames(life_table_m) <- c("Year", "Age", "mx", "qx", "ax", "lx", "dx", "Lx", "Tx", "ex")
male_2021 <- life_table_m[life_table_m$Year == 2021, ]
year_m <- male_2021$Year
age_m <- male_2021$Age
qx_m <- male_2021$qx
mx_m <- male_2021$mx
ax_m <- male_2021$ax
lx_m <- male_2021$lx
dx_m <- male_2021$dx
Lx_m <- male_2021$Lx
Tx_m <- male_2021$Tx
ex_m <- male_2021$ex

#female data
life_table_f <- read.table("C:\\Users\\megacrazyleo\\Desktop\\SQL\\R\\fltper_1x1.txt", header=FALSE, skip = 3, sep = "") 
colnames(life_table_f) <- c("Year", "Age", "mx", "qx", "ax", "lx", "dx", "Lx", "Tx", "ex")
female_2021 <- life_table_f[life_table_f$Year == 2021, ]
year_f <- female_2021$Year
age_f <- female_2021$Age
qx_f <- female_2021$qx
mx_f <- female_2021$mx
ax_f <- female_2021$ax
lx_f <- female_2021$lx
dx_f <- female_2021$dx
Lx_f <- female_2021$Lx
Tx_f <- female_2021$Tx
ex_f <- female_2021$ex
```

### Plot male vs female mortality rates in the year 2021
```{r}
plot(age_m, qx_m, 
     main = "Mortality rates (Canada, 2021)", 
     xlab = "Age x", 
     ylab = expression(paste("Mortality rate ", q[x])), 
     type = "l",
     col = "blue",
     )  # Using log scale for y-axis to better show differences

# Add the female line to the same plot
lines(age_f, qx_f, col = "red")


legend("bottomright", 
       legend = c("Male", "Female"),
       col = c("blue", "red"),
       lty = 1,
       cex = 0.8)

```
```{r}

# Create the plot with male data
plot(age_m, qx_m, 
    main = "Mortality rates (Canadian Males and Females, 2021)", 
    xlab = "Age x", 
    ylab = expression(paste("Mortality rate ", q[x])), 
    type = "l",
    col = "blue",
    log = "y",  
    ylim = range(c(qx_m, qx_f))  # Set y-axis limits to include both male and female data
)

# Add female data
lines(age_f, qx_f, col = "red")


grid()

# Add points to show actual data points
points(age_m, qx_m, pch = 20, cex = 0.5, col = "blue")
points(age_f, qx_f, pch = 20, cex = 0.5, col = "red")


legend("bottomright", 
       legend = c("Males", "Females"),
       col = c("blue", "red"),
       lty = 1,
       pch = 20,
       cex = 0.8)
```

### Find the probability for a 0 year old female and male in Canada to reach the age of 100 in 2021
```{r}
male_prob <- lx_m[100 + 1] / lx_m[0 + 1]
print(paste("Male probability:", male_prob))


female_prob <- lx_f[100 + 1] / lx_f[0 + 1]
print(paste("Female probability:", female_prob))
```
### Find the probability for a 0 year old in Canada to reach the age of 100 in 2021
```{r}
canada_prob <- lx[100 + 1] / lx[0 + 1]
print(paste("M/F probability:", canada_prob))
```
If we compare this with an 18 year old individual
```{r}
canada_prob_2 <- lx[100 + 1] / lx[18 + 1]
print(paste("M/F age 18 probability:", canada_prob_2))
```
Vs a 40 year old individual
```{r}
canada_prob_5 <- lx[100 + 1] / lx[40 + 1]
print(paste("M/F age 40 probability:", canada_prob_5))
```



And again with a 75 year old individual
```{r}
canada_prob_3 <- lx[100 + 1] / lx[75 + 1]
print(paste("M/F age 75 probability:", canada_prob_3))
```
Finally, compare with a 98 year old individual
```{r}
canada_prob_4 <- lx[100 + 1] / lx[98 + 1]
print(paste("M/F age 98 probability:", canada_prob_4))
```

```{r}
# Calculate probabilities
ages <- c(0, 18, 75, 98)
probs <- c(
  lx[100 + 1] / lx[0 + 1],
  lx[100 + 1] / lx[18 + 1],
  lx[100 + 1] / lx[40 + 1],
  lx[100 + 1] / lx[75 + 1],
  lx[100 + 1] / lx[98 + 1]
)

# Create a data frame
prob_data <- data.frame(Age = c("0", "18", "40", "75", "98"), Probability = probs)

# Create the bar plot
barplot(prob_data$Probability, 
        names.arg = prob_data$Age,
        main = "Probability of Reaching Age 100 from Different Starting Ages",
        xlab = "Starting Age",
        ylab = "Probability",
        col = "skyblue",
        ylim = c(0, max(probs) * 1.1))  # Set y-axis limit to 110% of max probability

# Add value labels on top of each bar
text(x = 1:5, 
     y = probs,
     labels = round(probs, 5),
     pos = 3,
     cex = 0.8)

# Add a grid for better readability
grid()

```

### Plot the survival probabilties for (18) up to age 100
```{r}
k <- 0:82
plot(k, lx[18 + k + 1] / lx[18 + 1], 
    pch = 20, 
    xlab = "k", 
    ylab = expression(paste(""[k], "p"[18])), 
    main = "Survival probabilities for age 18 up to 100")
```
Compare this with a 50 year old

```{r}
k <- 0:100
plot(k, lx[50 + k + 1] / lx[50 + 1], 
    pch = 20, 
    xlab = "k", 
    ylab = expression(paste(""[k], "p"[50])), 
    main = "Survival probabilities for age 50 up to 100")
```

### Plot the number of deaths
```{r}
# Plot the number of deaths dx by age
plot(age, dx, 
    type = "h", 
    pch = 20, 
    xlab = "Age x", 
    ylab = expression("d"[x]),
    main = "Number of deaths (Canada, M/F, 2021)")

# Simulate the number of deaths using a binomial distribution
sims <- rbinom(n = length(lx), size = lx, prob = qx)
  
# Plot the simulated number of deaths on top of the previous graph
points(age, sims, 
    pch = 4, 
    col = "red")

```




### Calculate the probability that (25) survives 5 more years
```{r}
prod(px[(25 + 1):(30 + 1)])
```

### Compute the survival probabilities of (25) until the age of 100

```{r}
kpx <- cumprod(px[(25 + 1):(99 + 1)])
```

### Extract the probability that (25) survives until the age of 100

```{r}
kpx[length(kpx)]
```

### Plot the probabilties for (25) to reach the age of 26, 27, ..., 100
```{r}
plot(1:length(kpx), kpx,
    pch = 20,
    xlab = "k",
    ylab = expression(paste(""[k], "p"[25])),
    main = "Survival probabilities for (25)")
```

### Compute the survival probabilities of (25)
```{r}
kpx <- c(1, cumprod(px[(25 + 1):(length(px) - 1)]))
```

### Compute the deferred mortality probabilities of (25)
```{r}
kqx <- kpx * qx[(25 + 1):length(qx)]
```


### Plot the deferred mortality probabilities of (25)
```{r}
plot(0:(length(kqx) - 1), kqx,
    pch = 20,
    xlab = "k",
    ylab = expression(paste(""['k|'], "q"[25])),
    main = "Deferred mortality probabilities of (25)")
```

## Life Expectancy


### Survival probabilities and curtate expected future lifetime of (0)
```{r}
kp0 <- cumprod(px)
sum(kp0)
```

### Survival probabilities and curtate expected future lifetime of a 25 year old
```{r}
kp25 <- cumprod(px[(25 + 1):length(px)])
sum(kp25)
```

### Complete expected future lifetime of (0) and (25)
```{r}
ex[c(0 + 1, 25 + 1)]
```

### Function to compute the curtate expected future lifetime for a given age and life table
```{r}
age2 <- as.numeric(age)
curtate_future_lifetime <- function(age2, life_table) {
  qx <- life_table$qx
  px <- 1 - qx
  kpx <- cumprod(px[(age2 + 1):length(px)])
  sum(kpx)
}
```


### Curtate future lifetimes for all ages
```{r}
valid_ages <- age2[!is.na(age2)]
future_lifetimes <- sapply(valid_ages, function(age) curtate_future_lifetime(age, life_table))
future_lifetimes
```

```{r}
# Future lifetime by age
plot(age2[!is.na(age2)], future_lifetimes, type = 'l', lwd = 2, col = "green", xlab = "Age x", ylab = "Future lifetime", main = "Future lifetime by age")
```


### Plot the logarithm of the male/female mortality rates for (18) by year
```{r}
# Load the life table
life_table <- read.table("C:\\Users\\megacrazyleo\\Desktop\\SQL\\R\\bltper_1x1.txt", header=FALSE, skip = 3, sep = "") 
colnames(life_table) <- c("Year", "Age", "mx", "qx", "ax", "lx", "dx", "Lx", "Tx", "ex")

# Ensure Age is numeric
life_table$Age <- as.numeric(life_table$Age)

# Subset the life table for age 18
subset_18 <- subset(life_table, Age == 18)

# Check if the subset has the same length for 'Year' and 'qx'
if (nrow(subset_18) > 0) {
  # Plot the log mortality rates for age 18
  with(subset_18,
       plot(Year, log(qx),
            type = "l", main = "Log mortality rates (Canada, M/F, 18-year-old)",
            xlab = "Year t", ylab = expression(paste("Log mortality rate ", log(q[18])))))
} else {
  print("No data available for age 18")
}

```



### Plot the logarithm of the female mortality rates in the year 1950 by age
```{r}
# Subset the life table for the year 1950
subset_1950 <- subset(life_table, Year == 1950)

# Check if the subset has the same length for 'Age' and 'qx'
if (nrow(subset_1950) > 0) {
  # Plot the log mortality rates for the year 1950
  with(subset_1950,
       plot(Age, log(qx),
            type = "l", main = "Log mortality rates (Canada, M/F, 1950)",
            xlab = "Age x", ylab = expression(paste("Log mortality rate ", log(q[x])))))
} else {
  print("No data available for the year 1950")
}
```

### Construct and print the cohort life table for birth year 1981
```{r}
life_table_1981 <- subset(life_table, Year - Age == 1981)
life_table_1981
```

### 1981 cohort one-year survival probabilities
```{r}
px <- 1 - life_table_1981$qx
```

### 1981 cohort survival probability that (18) survives 5 more years
```{r}
prod(px[(18 + 1):(22 + 1)])
```

### 1881 cohort survival probability that (18) survives 5 more years
```{r}
with(subset(life_table, Year - Age == 1881), prod(1 - qx[(18 + 1):(22 + 1)]))
```


### PV of guaranteed payment of 10,000 in 5 years
```{r}
PV <- 10000 * (1 + 0.02) ^ -5
PV
```

### 5 year survival probabilities of (20)
```{r}
kpx <- prod(px[(20 + 1):(24 + 1)])
```

### EPV of pure endowment of 10,000 in 5 years for (20)
```{r}
PV * kpx
```

# Life Annuities

### PV of guaranteed payments of 10,000 in 5, 10 and 30 years
```{r}
PV <- 10000 * (1 + 0.02) ^ - c(5, 10, 30)
PV
```

### Survival probabilities of (20)
```{r}
kpx <- cumprod(px[(20 + 1):length(px)])
```

### EPV of pure endowments of 10,000 in 5, 10 and 30 years for (20)
```{r}
PV * kpx[c(5, 10, 30)]
```

#whole life annuity due of 35
```{r}
kpx <-
  c(1, cumprod(px[(35+1):length(px)]))
discount_facotrs <- (1+0.03)^-(0:(length(kpx)-1))
benefits <- rep(1,length(kpx))
sum(benefits*discount_factors*kpx)
```

#whole life immediate annuity of 35
```{r}
kpx <-
  c(1, cumprod(px[(35+1):length(px)]))
discount_facotrs <- (1+0.03)^-(1:(length(kpx)))
benefits <- rep(1,length(kpx))
sum(benefits*discount_factors*kpx)
```

# Function to compute the EPV of a whole life annuity due for a given age, interest rate i and life table
```{r}
life_annuity_due <- function(age, i, life_table) {
  px <- 1 - life_table$qx
  kpx <- c(1, cumprod(px[(age + 1):length(px)]))
  discount_factors <- (1 + i) ^ - (0:(length(kpx) - 1))
  sum(discount_factors * kpx)
}
```

# EPV of a whole life annuity due for (20) at interest rate 2% using life_table
```{r}
life_annuity_due(20, 0.02, life_table)
```

# EPV of a whole life annuity due for (20) at interest rate 5% and for (65) at interest rate 2% using life_table
```{r}
life_annuity_due(20, 0.05, life_table)
life_annuity_due(65, 0.02, life_table)
```

# EPV of a whole life annuity due for (20) at interest rate 2% using life_table
```{r}
life_annuity_due(20, 0.02, life_table)
```

# Function to compute the EPV of a whole life immediate annuity for a given age, interest rate i and life table
```{r}
life_immediate_annuity <- function(age, i, life_table) {
  px <- 1 - life_table$qx
  kpx <- cumprod(px[(age + 1):length(px)])
  discount_factors <- (1 + i) ^ - (1:length(kpx))
  sum(discount_factors * kpx)
}
```

# EPV of a whole life immediate annuity for (20) at interest rate 2% using life_table
```{r}
life_immediate_annuity(20, 0.02, life_table)
```

# EPV of a whole life annuity due for (20) at interest rate 2% using life_table
```{r}
life_annuity_due(20, 0.02, life_table)
```

# Function to compute the EPV of a temporary life annuity due for a given age, period of n years, interest rate i and life table
```{r}
temporary_life_annuity_due <- function(age, n, i, life_table) {
  px <- 1 - life_table$qx
  kpx <- c(1, cumprod(px[(age + 1):(age + n - 1)]))
  discount_factors <- (1 + i) ^ - (0:(n - 1))
  sum(discount_factors * kpx)
}
```

# EPV of a temporary life annuity due for (20) over 10 years at interest rate 2% using life_table
```{r}
temporary_life_annuity_due(20, 10, 0.02, life_table)
```

### Calculating the PV of a pension at age 65
```{r}
benefits <- 20000 * 1.02 ^ (0:35)


# Discount factors (to age 65)

discount_factors <- 1.04 ^ - (0:35)


# PV of pension at age 65

PV_65 <- sum(benefits * discount_factors)
PV_65


# PV of pension at age 20

PV_20 <- PV_65 * 1.03 ^ - 45
PV_20
```


###  EPV of pension at age 20
# this might need some fixing
```{r}
# Survival probabilities of (65) up to age 100
px <- 1 - life_table$qx
kpx <- c(1, cumprod(px[(65 + 1):(99 + 1)]))


# EPV of pension at age 65

EPV_65 <- sum(benefits * discount_factors * kpx)
cbind(PV_65, EPV_65)


# EPV of pension at age 20

EPV_20 <- EPV_65 * (1.03 ^ - 45 * prod(px[(20 + 1):(64 + 1)]))
cbind(PV_20, EPV_20)
```

### Retirement Plan

Suppose 
# Survival probabilities of (40)
```{r}
kpx <- c(1, cumprod(px[(40 + 1):length(px)]))
```

# Discount factors (to age 40)
```{r}
discount_factors <- (1 + 0.03) ^ -(0:(length(kpx) - 1))
```

# Pension benefits
```{r}
benefits <- c(rep(0, 25), rep(18000, length(kpx) - 25))
```

# The single premium
```{r}
single_premium <- sum(benefits * discount_factors * kpx)
single_premium
```

# Premium pattern rho
```{r}
rho <- c(rep(1, 15), rep(0.5, 10), rep(0, length(kpx) - 25))
```

# The initial premium
```{r}
initial_premium <- single_premium / sum(rho * discount_factors * kpx)
initial_premium
```

# The annual premiums
initial_premium * rho

# Sum of the annual premiums (no actuarial discounting)
sum(initial_premium * rho)

# Curtate life expectancy of (40)
sum(kpx[-1])

# Present value of annuity benefits when (40) lives until age 75
subset1 <- 1:36
sum(benefits[subset1] * discount_factors[subset1])

# Present value of annuity benefits when (40) lives until age 95
subset2 <- 1:56
sum(benefits[subset2] * discount_factors[subset2])

# 10-year survival probability of (20)
kpx <- prod(px[(20 + 1):(29 + 1)])
kpx

# 10-year deferred mortality probability of (20)
kqx <- kpx * qx[30 + 1]
kqx

# Discount factor
discount_factor <- (1 + 0.01) ^ - 11
discount_factor

# EPV of the simple life insurance
10000 * discount_factor * kqx

# Function to compute the EPV of a whole life insurance
whole_life_insurance <- function(age, i, life_table) {
  qx <- life_table$qx
  px <- 1 - qx
  kpx <- c(1, cumprod(px[(age + 1):(length(px) - 1)]))
  kqx <- kpx * qx[(age + 1):length(qx)]
  discount_factors <- (1 + i) ^ - (1:length(kqx))
  sum(discount_factors * kqx)
}

# Plot the EPV of a whole life insurance for a range of ages at interest rate 3% using life_table
plot_by_age()

# Plot the EPV of a whole life insurance for (20) for a range of interest rates using life_table
plot_by_interest_rate()

# EPV of a whole life insurance for (20) at interest rate 2% using life_table
whole_life_insurance(20, 0.02, life_table)

# Function to compute the EPV of a temporary life insurance
temporary_life_insurance <- function(age, n, i, life_table) {
  qx <- life_table$qx
  px <- 1 - qx
  kpx <- c(1, cumprod(px[(age + 1):(age + n - 1)]))
  kqx <- kpx * qx[(age + 1):(age + n)]
  discount_factors <- (1 + i) ^ - (1:length(kqx))
  sum(discount_factors * kqx)
}

# EPV of a temporary life insurance for (20) over a period of 45 years at interest rate 2% using life_table
temporary_life_insurance(20, 45, 0.02, life_table)

# EPV of a whole life insurance for (20) at interest rate 2% using life_table
whole_life_insurance(20, 0.02, life_table)

# Function to compute the EPV of a deferred whole life insurance
deferred_life_insurance <- function(age, u, i, life_table) {
  qx <- life_table$qx;  px <- 1 - qx
  kpx <- c(1, cumprod(px[(age + 1):(length(px) - 1)]))
  kqx <- kpx * qx[(age + 1):length(qx)]
  discount_factors <- (1 + i) ^ - (1:length(kqx))
  benefits <- c(rep(0, u), rep(1, length(kpx) - u))
  sum(benefits * discount_factors * kqx)
}

# EPV of a deferred life insurance for (20) deferred over 45 years at interest rate 2% using life_table
deferred_life_insurance(20, 45, 0.02, life_table)

# Deferred mortality probabilites of (48)
kqx <- c(1, cumprod(px[(48 + 1):(73 + 1)])) * qx[(48 + 1):(74 + 1)]

# Discount factors
discount_factors <- (1 + i) ^ - (1:length(kqx))

# Death benefits
benefits <- c(rep(0, 7), rep(40000, length(kqx) - 7))

# EPV of the death benefits
EPV_death_benefits <- sum(benefits * discount_factors * kqx)
EPV_death_benefits

# Pure endowment
EPV_pure_endowment <- 80000 * (1 + i) ^ - 27 * prod(px[(48 + 1):(74 + 1)])
EPV_pure_endowment

# Premium pattern
kpx <- c(1, cumprod(px[(48 + 1):(73 + 1)]))
discount_factors <- (1 + i) ^ - (0:(length(kpx) - 1))
rho <- rep(1, length(kpx))
EPV_rho <- sum(rho * discount_factors * kpx)
EPV_rho

# Premium level
(EPV_death_benefits + EPV_pure_endowment) / EPV_rho


