---
title: "Life Insurance Analysis"
author: "Leo Cam 
[My Github](https://github.com/cam-leo/index.html)"
date: "2023-01-22"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    theme: united
    highlight: tango
    code_folding: hide
    css: combined.css
    self_contained: false
---
<div class="content-wrapper">

```{r install-packages, include=FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com/"))
packages <- c("shiny", "dplyr", "forecast", "mgcv", "survival")
for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "figures/")
# Load required libraries
library(shiny)
library(dplyr)

# ---- Data loading -------------------------------------------------------
# Life tables come from the Human Mortality Database (https://mortality.org),
# Canada, period 1x1. Put the raw HMD files in a `data/` folder next to this
# file. If the combined (both sexes) file isn't there, the bundled
# life_table.rds (same data, already cleaned) is used instead.
read_hmd <- function(file) {
  lt <- read.table(file, header = TRUE, skip = 2, stringsAsFactors = FALSE)
  colnames(lt) <- c("Year", "Age", "mx", "qx", "ax", "lx", "dx", "Lx", "Tx", "ex")
  lt$Age <- as.numeric(sub("+", "", lt$Age, fixed = TRUE))  # "110+" -> 110
  lt[] <- lapply(lt, as.numeric)
  lt
}

load_life_table <- function(sex = c("both", "male", "female")) {
  sex <- match.arg(sex)
  file <- file.path("data", c(both = "bltper_1x1.txt", male = "mltper_1x1.txt",
                              female = "fltper_1x1.txt")[[sex]])
  if (file.exists(file)) return(read_hmd(file))
  if (sex == "both" && file.exists("life_table.rds")) {
    lt <- readRDS("life_table.rds")
    lt$Age[is.na(lt$Age)] <- 110  # the open age group "110+" was stored as NA
    return(lt)
  }
  NULL
}

life_table_all <- load_life_table("both")
if (is.null(life_table_all)) stop("No life table found: add data/bltper_1x1.txt or life_table.rds")
life_table_2021 <- subset(life_table_all, Year == 2021)

life_table_m_all <- load_life_table("male")
life_table_f_all <- load_life_table("female")
has_sex_tables <- !is.null(life_table_m_all) && !is.null(life_table_f_all)
```

---------------------------------------------------------------------------------

# Introduction


This document is divided into four parts, each focusing on different aspects of financial and actuarial analysis<sup><a href="#" title="Actuarial analysis involves the application of mathematical and statistical methods to assess risk in insurance, finance, and other industries. Actuaries use this analysis to model and predict future events, such as life expectancy, accident frequency, and financial losses, enabling companies to develop policies and strategies to mitigate potential risks. This process includes evaluating the likelihood of events, estimating their financial impact, and using this information to set premiums, reserves, and other financial metrics. Actuarial analysis is essential for ensuring the financial stability and sustainability of insurance companies and pension plans.">(?)</a></sup> related to life insurance<sup><a href="#" title="Life insurance is a contract between an individual and an insurance company. The individual pays regular premiums, and in return, the insurance company promises to pay a specified amount of money to designated beneficiaries upon the insured person's death. This financial protection helps beneficiaries cover expenses such as funeral costs, debts, and living expenses, ensuring financial stability during a difficult time.">(?)</a></sup>. I am using actuarial analysis because there is lots of data that we can learn and use from actuaries in overall data analysis. For most key subjects, there is a small question mark that you can hover over and it will give you a blurb of information describing what that topic is. ***If you want to only see the calculations, you can skip ahead through the TOC on the left and click on 'show' to see the code***:

-----------------------------------------------------------------------------------

**Cash Flows**

***Topic***: *Present Value (PV)<sup><a href="#" title="Present Value (PV) represents the current worth of a future sum of money, discounted back to the present using an appropriate discount rate. This concept is rooted in probability theory, where future cash flows are uncertain and thus discounted to reflect the time value of money and the associated risk. Statistical inference techniques help determine appropriate discount rates by analyzing historical data and estimating future uncertainties, such as inflation rates or market risks.">(?)</a></sup>, Net Present Value (NPV)<sup><a href="#" title="Net Present Value (NPV) builds upon PV by considering the difference between the present value of cash inflows and outflows over a specified period. NPV incorporates probability theory by assessing the likelihood of achieving expected cash flows, often using statistical methods to model uncertain future outcomes. It serves as a critical decision-making tool, indicating whether an investment or project is expected to generate positive returns after accounting for the time value of money and risk.">(?)</a></sup>, Accumulated Value (AV)<sup><a href="#" title="Accumulated Value (AV) refers to the total value of an investment or project at a specific point in time, taking into account all cash flows, including initial investments, returns, and reinvestments. Probability theory assists in forecasting AV by analyzing the distribution of possible future cash flows and their associated uncertainties. Statistical inference techniques, such as Monte Carlo simulations or sensitivity analysis, further refine these forecasts by exploring different scenarios and their impacts on AV over time.">(?)</a></sup>, Statistical Inference<sup><a href="#" title="Statistical inference uses data from a sample to make guesses about a larger population. It helps us draw conclusions and make decisions based on limited information.">(?)</a></sup> and Probability Theory<sup><a href="#" title="Probability theory is the study of chance. It helps us understand how likely events are to happen, like rolling a die or predicting the weather.">(?)</a></sup>*

Cash flows are fundamental to financial analysis, involving the movement of money into (income) and out of (expenses) a business or project over time. The concepts of <span style="color: blue;">Present value (PV)</span>, <span style="color: blue;">Net Present Value (NPV)</span>, and <span style="color: blue;">Accumulated Value (AV)</span> are crucial tools in evaluating cash flows, as they can provide insights into their current and future worth.

In summary, integrating statistical inference and probability theory enhances the analysis of cash flows by providing methods to quantify uncertainties, estimate future values, and assess risks. PV, NPV, and AV serve as powerful tools within this framework, enabling informed financial decisions by accounting for the complex interplay between time, risk, and value in financial management and investment evaluation.

**Life Tables**

***Topic***: *Survival probabilities<sup><a href="#" title="Survival probabilities refer to the likelihood of individuals surviving from one age to the next within a specified population or cohort. These probabilities are fundamental in demography, actuarial science, and medical research, where they are used to analyze and predict how long individuals are expected to live.">(?)</a></sup>, Probability and Statistics*

Life tables are foundational in demography and actuarial science, providing a statistical framework to analyze and predict survival probabilities within a population. These tables import and organize mortality data, offering insights into how longevity varies across different age groups and over time.

Life tables utilize probability theory to estimate mortality probabilities (qx) at various ages, the likelihood that someone aged x dies before x + 1. Survival probabilities are their complement, px = 1 - qx. Both are derived from historical mortality data. Statistical inference techniques are employed to analyze and interpret these probabilities, including:

*Estimation Methods*: Statistical methods such as <span style="color: blue;">maximum likelihood estimation</span><sup><a href="#" title="Maximum Likelihood Estimation (MLE) is a method used to estimate the parameters of a statistical model by maximizing the likelihood function. The likelihood function represents the probability of observing the given data under a specific set of parameter values. MLE selects the parameter values that make the observed data most likely, assuming the model is correct. It is widely used in statistical modeling and inference to find the best-fitting model parameters based on the available data.">(?)</a></sup> (MLE) or <span style="color: blue;">Bayesian inference</span><sup><a href="#" title="Bayesian inference is a statistical approach that combines prior beliefs or knowledge with observed data to update and quantify uncertainty about a hypothesis or parameter of interest. Unlike frequentist statistics, which focuses on estimating parameters based solely on observed data, Bayesian inference incorporates prior information into the analysis. It provides a framework for updating beliefs as new evidence becomes available, making it especially useful in decision-making under uncertainty.">(?)</a></sup> are used to derive mortality probabilities (qx) from observed central death rates (mx)<sup><a href="#" title="In demography and actuarial science, mx represents the central mortality rate at exact age x within a population. It is a rate rather than a probability: deaths at age x divided by the person-years lived at age x (the population exposed to risk). mx values are essential in constructing life tables, survival models, and mortality projections. They provide insights into age-specific mortality patterns and are used to calculate life expectancies, assess insurance risks, and plan for retirement and healthcare needs..">(?)</a></sup>.

*Population Dynamics*: Statistical analysis of life tables informs demographic studies, healthcare planning, and insurance risk assessment by predicting population growth, mortality patterns, and aging trends.

In summary, probability and statistics play a critical role in understanding life tables and mortality data sets. They facilitate the analysis of survival probabilities (qx) and provide valuable insights into population dynamics, longevity trends, and risk assessment in various fields, from public health to insurance and pension planning.


**Life Annuities**

***Topic***: *Generalized Linear Models (GLMs)<sup><a href="#" title="GLMs are a way to predict outcomes using data. They extend regular linear models to handle different types of data, like counts or categories.">(?)</a></sup> and Time Series Analysis<sup><a href="#" title="Time series analysis looks at data collected over time. It helps us find patterns, like trends or cycles, and make forecasts about the future.">(?)</a></sup>*

Life annuities are financial products that provide a stream of payments to an individual over their lifetime or for a specified period. Analyzing these annuities involves applying statistical methods like <span style="color: blue;">Generalized Linear Models (GLMs)</span> and <span style="color: blue;">Time Series Analysis</span> to estimate their PV and expected future payments.

GLMs are utilized to model the relationship between factors like age, gender and economic conditions with the PV of an annuity. They help adjust for non-linear relationships and <span style="color: blue;">heteroscedasticity</span><sup><a href="#" title="Heteroscedasticity refers to the unequal variance of errors (or residuals) in a regression model. In simpler terms, it indicates that the variability of the dependent variable (y) is not constant across all levels of the independent variable(s). This violates the assumption of homoscedasticity, where the variability of y is consistent. Heteroscedasticity can affect the accuracy of statistical tests and predictions, requiring adjustments in modeling techniques to account for varying levels of variability.">(?)</a></sup> in annuity pricing. GLMs also allow for the use of risk factors like mortality rates <span style="color: blue;">(qx)</span><sup><a href="#" title="In actuarial science and demography, qx represents the probability of dying within a specified age interval, often referred to as the mortality rate at age x. It is a fundamental measure used in life tables and survival models to analyze and predict the likelihood of death at different ages within a population. qx values are derived from mortality data and are crucial for assessing life expectancy, pricing life insurance products, and evaluating healthcare and pension plans.">(?)</a></sup>, interest rates and such into pricing models. This way, insurers can assess and price the risk associated with annuity payments.

Time Series Analysis is another critical tool for analyzing annuities, particularly for forecasting future payments and assessing long-term trends:

Models like <span style="color: blue;">ARIMA</span><sup><a href="#" title="ARIMA is a statistical model used for time series analysis and forecasting. It combines autoregressive (AR), moving average (MA), and differencing components (I for Integrated) to model the temporal structure and patterns in data. ARIMA models are useful for predicting future values based on past observations and are widely applied in economics, finance, and other fields where time-dependent data is analyzed.">(?)</a></sup> (*AutoRegressive Integrated Moving Average*) or exponential smooth models can be used to forecast future cash flows from annuities. The Expected Present Value (EPV) is calculated as the discounted sum of expected future payments, incorporating forecasts generated from time series models. Time series also allow for stochastic modeling of uncertain factors affecting annuity payments, such as mortality improvements and economic fluctuations. One such model are the <span style="color: blue;">Monte Carlo simulations</span><sup><a href="#" title="Computational techniques used to understand the impact of uncertainty and variability in mathematical, financial, and scientific models. They involve running numerous random simulations based on input variables that have uncertain values. By repeating these simulations many times, Monte Carlo methods provide estimates of possible outcomes and their probabilities.">(?)</a></sup>, which provide probabilistic forecasts of future annuity cash flows

Application in Financial Decision-Making:

The integration of GLMs and Time Series Analysis enhances the understanding and management of life annuities. In product design, statistical models inform the design of annuity products tailored to demographic trends and customer preferences, ensuring competitive pricing. In risk management, by quantifying risks and uncertainties through GLMs and Time Series Analysis, insurers and financial planners can optimize portfolio allocation and hedging strategies to mitigate adverse impacts on annuity liabilities

In conclusion, GLMs and Time Series Analysis are powerful tools in the evaluation and management of life annuities. They facilitate the calculation of present values (PV) and expected future payments (EPV), providing insights into pricing dynamics, risk assessment, and strategic decision-making in insurance and retirement planning.

**Life Insurance**

***Topic***: *Survival Models<sup><a href="#" title="Survival models predict the time until an event occurs, like how long a patient might live after treatment. They help in planning and understanding risks.">(?)</a></sup> and Credibility Theory<sup><a href="#" title="Credibility theory combines past data with new data to make better predictions. It's often used in insurance to set fair premiums based on individual risk.">(?)</a></sup>*

Life insurance products provide financial protection against the risk of premature death, offering payouts (benefits) to beneficiaries upon the death of the insured. Analyzing these products involves applying statistical methods such as Survival Models and Credibility Theory to assess premiums and estimate their expected future values.

<span style="color: blue;">Survival Models</span><sup><a href="#" title="Survival models are statistical tools used to analyze and predict the time until an event of interest occurs, such as death, failure of a device, or recurrence of a disease. These models help understand factors influencing survival times and can account for censored data (where the event of interest hasn't occurred for all subjects by the end of the study).">(?)</a></sup>, such as the <span style="color: blue;">Kaplan-Meier</span><sup><a href="#" title="The Kaplan-Meier estimator is a statistical method used to estimate the probability of survival over time. It's often used in medical and biological studies to analyze how long patients survive after treatment or diagnosis. It calculates survival probabilities based on observed data, especially when not all individuals have experienced the event of interest (such as death) by the end of the study.">(?)</a></sup> estimator or <span style="color: blue;">Cox proportional hazards model</span><sup><a href="#" title="The Cox proportional hazards model is a statistical technique used to assess the influence of multiple factors (like age, gender, and treatment) on the timing of a specific event (such as death). It doesn't assume a specific distribution for the survival times and allows for the estimation of hazard ratios, which describe how the hazard (or risk) of an event changes with different variables.">(?)</a></sup>, are statistical techniques used to model and analyze the time until an event occurs—in this case, the death of the insured. In the context of life insurance:

Survival models can estimate mortality rates (qx) and predict the likelihood of death at different ages and for different demographic groups. These models help the calculation of premiums by assessing the risk of the insured dying before the policy term. They also help quanity the risk associated with these life insurance policies by considering numerous factors like age, gender, health status etc. By modeling mortality risks, insurers can set appropriate premiums that reflect the expected costs of providing insurance coverage.

<span style="color: blue;">Credibility Theory</span><sup><a href="#" title="Credibility theory is a statistical framework used in insurance and risk management to improve the accuracy of predictions when data is limited or variable. It combines historical data with prior beliefs or external information to estimate future outcomes more reliably. In insurance, it helps set appropriate premiums by adjusting for uncertainties and variations in risk across different policyholders or groups.">(?)</a></sup> is a statistical framework used to improve the accuracy of risk estimation when data is limited or variable. Key aspects in the context of life insurance include:

It can enhance the estimation of Expected Present Value (EPV) through combining historical data with prior beliefs / external info (credibility factors), as well as help insurers reduce estimation errors by incorporating these credibility factors when they optimize pricing strategies

*Application in Insurance Planning*:

The integration of Survival Models and Credibility Theory supports strategic decision-making in life insurance. Statistical models can inform the design of life insurance products tailored to customer needs and risk profiles, as well as provide insights that enable insurers to offer competitive products. Bu helping insurers assess mortality risks, insurers can manage reinsurance arrangements and capital adequacy to mitigate adverse financial impacts from unexpected claims

In summary, Survival Models and Credibility Theory are essential tools in the analysis and management of life insurance. They facilitate the calculation of premiums, estimation of EPV, and strategic decision-making, ensuring sustainable risk management and competitive product offerings in the insurance industry.

---------------------------------------------------------------------------------

# Cash Flows

---------------------------------------------------------------------------------

## Present Value

<details>
  <summary>What is Present Value (PV)?</summary>
  <p style="font-weight: bold; color: blue;">
  Present value is the current worth of a future sum of money. It tells you how much money you would need to invest today to reach a specific amount in the future, considering a certain interest rate.
   </p>
</details>




Suppose you want to go on a trip 5 years from now. You have expenses of $3000 each year for 4 years (including this year), followed by an expense of $4000 and a final expense of $5000. How much would you have to deposit at this moment in time to cover all of these future payments? Assume a constant interest rate of 2.14% and payments would take place at the beginning of the year.


```{r}
# Define cash flows and time periods
cash_flows <- c(3000, 3000, 3000, 3000, 4000, 5000)
time_periods <- 0:6  

# Create the plot
plot(time_periods, rep(0, length(time_periods)), type = "n", 
     ylim = c(0, max(cash_flows) * 1.1),
     xlim = c(0, 6),  
     xlab = "t", ylab = "",
     main = "Cash Flow Timeline",
     xaxt = "n")


axis(1, at = time_periods)

# Here we add labels for the separate cash flows
for (i in 1:length(cash_flows)) {
  lines(c(i - 1, i - 1), c(0, cash_flows[i]))
  text(i - 1, cash_flows[i], labels = cash_flows[i], pos = 3)
}

# Add horizontal line to connect all labels together and provide some more cohesiveness
abline(h = 0)




```

Let the annual interest rate be \\(i = 0.0214\\).

The discount factor \\(v\\) is calculated as follows:
\\[
v = \\frac{1}{1+i}
\\]


<details>
  <summary>What is discount factor?</summary>
  <p style="font-weight: bold; color: blue;">
  A discount factor is a number used to calculate the present value of a future amount of money. It shows how much a future sum is worth today, considering a specific interest rate over time.
  </p>
</details>




The discount factors for the time periods \\(t = 0\\) to \\(t = 5\\) are given by:
\\[
v^{(0:5)} = \\left(v^0, v^1, v^2, v^3, v^4, v^5\\right)
\\]

The present value (PV) of the cash flows is calculated as the sum of the cash flows multiplied by the corresponding discount factors:
\\[
\\text{PV} = \\sum_{t=0}^{5} (\\text{cash flow at } t) \\times v^t
\\]

```{r}
# We have an expense of 3000 from t=0 to t=3, an expense of 4000 at t=4 and 5000 at t=5

# Cash flows
cash_flows <- c(rep(3000,4), 4000, 5000)
                
# Interest Rate
i <- 0.0214

# Discount factor
v <- 1 / (1+i)

# Discount factors
discount_factors <- v ^ (0:5)

# Present value PV
present_val <- sum(cash_flows * discount_factors)
cat("Present Value (PV):", present_val)
```

So you would have to make a deposit of 19800.98 at this current time to cover all of these future payments.

##### How sensitive is the deposit to the interest rate?

These six payments are fixed and known, so the PV of `r format(round(present_val, 2), nsmall = 2)` is an exact number rather than an estimate. There is no sample to bootstrap and nothing to test: resampling the payments with replacement (as an earlier version did) shuffles which amount is paid when and doesn't describe any real uncertainty.

What *is* uncertain is the interest rate you will actually earn. Below, the rate is assumed to be normally distributed around 2.14% with a standard deviation of 0.5% (an illustrative assumption), and we look at how the required deposit varies and how likely a round $20,000 deposit is to be enough.

```{r}
set.seed(123)
n_sim <- 10000
rates <- rnorm(n_sim, mean = 0.0214, sd = 0.005)
pv_sim <- sapply(rates, function(r) sum(cash_flows * (1 + r) ^ -(0:5)))

cat("Middle 95% of the required deposit:", round(quantile(pv_sim, c(0.025, 0.975)), 2), "\n")
cat("Probability that a $20,000 deposit is enough:", mean(pv_sim <= 20000), "\n")

# Interest rate at which exactly $20,000 is needed
breakeven_rate <- uniroot(function(r) sum(cash_flows * (1 + r) ^ -(0:5)) - 20000, c(-0.05, 0.2))$root
cat("$20,000 is enough whenever the rate is at least", round(100 * breakeven_rate, 3), "%\n")

hist(pv_sim, breaks = 50, main = "Required deposit under interest-rate uncertainty",
     xlab = "Deposit needed today")
abline(v = c(present_val, 20000), col = c("red", "blue"), lwd = 2, lty = c(1, 2))
legend("topright", c("PV at 2.14%", "$20,000"), col = c("red", "blue"), lty = c(1, 2), lwd = 2)
```

A higher interest rate means less needs to be deposited today, so $20,000 covers the trip as long as the rate stays above about `r round(100 * breakeven_rate, 2)`%.


------------------------------------------------------------------------------------

## Net Present Value (NPV)

<details>
  <summary>What is Net Present value (NPV)?</summary>
  <p style="font-weight: bold; color: blue;">
  Net Present Value (NPV) is a financial metric that calculates the value of a series of future cash flows in today's terms. It takes into account the time value of money by discounting future cash flows back to the present. NPV helps determine the profitability of an investment: a positive NPV indicates a good investment, while a negative NPV suggests it may not be worthwhile.
   </p>
</details>

You're the head of a lawn mowing company. You want to invest in a new lawn mower than you are confident will generate cash
flows of 50 each month for 3 months, followed by 20 for 2 months. The issue is it costs 185 and you're unsure if it will be a profitable investment. To find out if it is a good investment or not, you must calculate the NPV for this investment. Assume the monthly interest rate is 0.6%.


```{r}

# Define the cash flows
cash_flows <- c(-185, rep(50, 3), rep(20, 2))

# Discount factors
monthly_interest_rate <- 0.006
discount_factors <- (1 + monthly_interest_rate) ^ -(0:5)

times <- 0:5


plot(times, cash_flows, type="n", xlab="Time (t)", ylab="Cash Flows", main="Cash Flow Timeline", ylim=c(-200, 70))


abline(h = 0)


segments(times, 0, times, cash_flows)

# Add text labels for the cash flows
text(times, cash_flows, labels=cash_flows, pos=ifelse(cash_flows < 0, 1, 3))


axis(1, at=times, labels=times)
axis(2, labels=FALSE)

# NPV calculation
npv <- sum(cash_flows * discount_factors)
cat("Net Present Value (NPV):", npv, "\n")
```

The Net Present Value (NPV) is calculated using the formula:
\\[
\\text{NPV} = -C_0 + \\sum_{t=1}^{n} CF_t \\, v^t, \\qquad v = \\frac{1}{1+r}
\\]


Where:

- \\( CF_t \\) is the cash flow at time \\( t \\)

- \\( r \\) is the monthly interest rate

- \\( C_0 \\) is the initial investment cost

- \\( n \\) is the number of periods

Given:

- Cash flows: \\( CF_1 = 50, CF_2 = 50, CF_3 = 50, CF_4 = 20, CF_5 = 20 \\)

- Initial investment: \\( C_0 = 185 \\)

- Monthly interest rate: \\( r = 0.006 \\)

Since the NPV is positive (`r round(npv, 2)`), the mower is expected to be profitable, but only just: the margin is about 1% of the purchase price.

To better understand the risk of the investment, we can simulate the future cash flows under uncertainty<sup><a href="#" title="Uncertainty means not knowing exactly what will happen in the future. In this context, it refers to the unpredictable factors that can affect the cash flows of an investment, such as market fluctuations or economic changes.">(?)</a></sup>. The purchase price is known, so only the monthly earnings vary; we assume each month's earnings are off by a normally distributed amount with a standard deviation of 5 (an illustrative assumption).

```{r}
# Simulation for NPV under uncertainty
set.seed(123)
n_simulations <- 1000
simulated_npvs <- replicate(n_simulations, {
  # the price at t = 0 is fixed; only future earnings are uncertain
  simulated_cash_flows <- cash_flows + c(0, rnorm(length(cash_flows) - 1, mean = 0, sd = 5))
  sum(simulated_cash_flows * discount_factors)
})

# Calculate the mean and standard deviation of simulated NPVs
mean_npv <- mean(simulated_npvs)
sd_npv <- sd(simulated_npvs)
cat("Mean NPV:", mean_npv, "\n")
cat("Standard Deviation of NPV:", sd_npv, "\n")
cat("Probability the investment loses money (NPV < 0):", mean(simulated_npvs < 0), "\n")

# Plot the distribution of simulated NPVs
hist(simulated_npvs, breaks = 30, main = "Distribution of Simulated NPVs", xlab = "NPV")
abline(v = mean_npv, col = "red", lwd = 2)
abline(v = 0, lty = 2)


```

Because the expected NPV is so close to zero, modest variation in monthly earnings makes a loss quite likely (about `r round(100 * mean(simulated_npvs < 0))`% of simulations), so this is a marginal investment.

------------------------------------------------------------------------------------

## Accumulated Value

<details>
  <summary>What is Accumulated Value (AV)?</summary>
  <p style="font-weight: bold; color: blue;">
   The total amount of money accumulated or earned over time from an initial investment or savings, including both the original amount invested (principal) and any interest or earnings gained on it.
   </p>
</details>

Suppose instead of making those expenses, you chose to deposit the same amounts into a savings account at the same times. What is the present value of those deposits, and how much would the account hold at t = 6 (the accumulated value)?

```{r}

# Cash flows
cash_flows <- c(rep(3000, 4), 4000, 5000)
cat("Cash flows:", cash_flows)

# Define the discount function 
discount <- function(s, t, i = 0.0214) {
  (1 + i) ^ - (t - s)
}

# Calculate the present value PV
pv <- sum(cash_flows * discount(0, 0:5))
cat("Present Value (PV):", pv)

# Calculate the value at time 6
av <- sum(cash_flows * discount(6, 0:5))
cat("Accumulated Value (AV):", av)

# Calculate the value at time 6, starting from the PV
pv_at_6 <- pv * discount(6, 0)
cat("PV at time 6:", pv_at_6)

interest_earned <- av - sum(cash_flows)
cat("Total deposited:", sum(cash_flows), "  Interest earned:", interest_earned)

```

Saving instead of spending leaves `r format(round(av, 2), nsmall = 2)` in the account at t = 6: the `r sum(cash_flows)` you deposited plus `r format(round(interest_earned, 2), nsmall = 2)` of interest. The PV is unchanged (the same payments at the same times), and accumulating the PV to t = 6 gives exactly the AV, which is a useful consistency check.

#### Sensitivity Analysis of Accumulated Value

To understand how sensitive the accumulated value is to changes in the interest rate, we can perform a sensitivity analysis<sup><a href="#" title="Sensitivity analysis helps us understand how changes in one factor, like sales or costs, affect outcomes such as profit or investment returns. It shows the impact of these changes, helping to make better decisions or prepare for different scenarios.">(?)</a></sup>
.

```{r}
# Sensitivity analysis for different interest rates
interest_rates <- seq(0.01, 0.05, by = 0.005)
accumulated_values <- sapply(interest_rates, function(i) {
  discount <- function(s, t) (1 + i) ^ - (t - s)
  sum(cash_flows * discount(6, 0:5))
})

# Plot the sensitivity analysis
plot(interest_rates, accumulated_values, type = "b", 
     xlab = "Interest Rate", ylab = "Accumulated Value", 
     main = "Sensitivity Analysis of Accumulated Value")
abline(v = 0.0214, col = "red", lwd = 2, lty = 2)
cat("Accumulated Values for different interest rates:", accumulated_values)

```


#### What interest rate would reach $25,000?

As with the PV, the AV is an exact number once the rate is fixed, so a bootstrap interval or hypothesis test around it isn't meaningful. A more useful question is what rate you would need for the savings to reach a target, say $25,000 at t = 6.

```{r}
target_av <- 25000
av_at <- function(r) sum(cash_flows * (1 + r) ^ (6 - 0:5))
required_rate <- uniroot(function(r) av_at(r) - target_av, c(0, 0.5))$root
cat("Interest rate needed for an AV of $25,000:", round(100 * required_rate, 3), "%\n")
cat("Check: AV at that rate =", round(av_at(required_rate), 2))
```

You would need about `r round(100 * required_rate, 2)`% a year, compared with the 2.14% assumed.


------------------------------------------------------------------------------------

## Yearly Payments

Your parents want to start saving money for your university costs. Each year of study will set you guys back by about 7500. You're currently 14 (t=0) and will start attending when you are 19 (t=5). Therefore, your parents will make 1 deposit each year for 4 years (t=1 to t=4). Assume you will attend for 4 years (t=5 to t=8), the constant annual interest rate is 3.14% and each deposit your parents make is equal. Find the amount of the deposit





```{r}

# discount factors
discount_factors <- (1 + 0.0314) ^ - (0:8)

# deposit pattern
deposits <- c(0, rep(1, 4), rep(0, 4))

# university expenses
payments <- c(rep(0, 5), rep(7500, 4))

# Calculate the PV of the deposits
PV_deposit <- sum(deposits * discount_factors)

# Calculate the PV of the payments
PV_payment <- sum(payments * discount_factors)

# Calculate the yearly deposit K in the first 4 years
yearly_deposit <-  PV_payment / PV_deposit 
cat("Yearly Payments =", yearly_deposit)

# Cash flow timeline: deposits in (+), tuition out (-)
times <- 0:8
cash_flows <- yearly_deposit * deposits - payments
plot(times, cash_flows, type = "n", xlab = "Time (t)", ylab = "Cash Flows",
     main = "Cash Flow Timeline", ylim = c(-9000, 9000), yaxt = "n")
abline(h = 0)
segments(times, 0, times, cash_flows)
nz <- cash_flows != 0
text(times[nz], cash_flows[nz], labels = round(cash_flows[nz], 2), pos = ifelse(cash_flows[nz] < 0, 1, 3))

```

So your parents would have to make a deposit of `r format(round(yearly_deposit, 2), nsmall = 2)` for each of the 4 years to finance your university.

------------------------------------------------------------------------------------

## Varying Interest Rates

You want to take out a loan to open a new restaurant. You plan to take $1000 this year and $3000 next year. You plan to repay the money with equal yearly payments for the 10 years after next year (t=2 to t=11). Note this time the interest rate isn’t constant but changes over time. For the first 3 years it is 4%, then for the next 3 years it is 4.5%, then for the final 5 years it is 6%. How much will your yearly payments be?

```{r, fig.width=12, fig.height=6}
# Interest rates
interest <- c(rep(0.04, 3), rep(0.045, 3), rep(0.06, 5))

# Yearly discount factors
yearly_discount_factors <- (1 + interest) ^ (-1)

# Discount factors
discount_factors <- cumprod(c(1, yearly_discount_factors))

# Cash flows for the first two years
cash_flow <- c(1000, 3000)

# Calculate the PV of the loan
PV_loan <- sum(cash_flow * discount_factors[1:2])

# Calculate the PV of the repayments
repayment_years <- 2:11
discount_factors_repayments <- discount_factors[repayment_years + 1]
n_payments <- length(discount_factors_repayments)


yearly_payment <- PV_loan / sum(discount_factors_repayments)


cash_flow_full <- c(cash_flow, rep(-yearly_payment, n_payments))

# Calculate the PV of the repayments to check the balance
PV_repayments <- sum(cash_flow_full * discount_factors)

# Display the yearly payment and the PV balance check
cat("Yearly payments =", yearly_payment, "\n")
cat("Net PV of loan minus repayments (should be 0) =", PV_repayments)

# Cash flow timeline using the computed payment
# borrower's view: loan received (+), repayments paid (-)
cash_flows <- c(1000, 3000, rep(-round(yearly_payment, 2), 10))
time_periods <- 0:12

# Calculate padding for y-axis
y_padding <- max(abs(cash_flows)) * 0.2




plot(time_periods, rep(0, length(time_periods)), type = "n", 
     ylim = c(min(cash_flows) - y_padding, max(cash_flows) + y_padding),
     xlim = c(0, 12),
     xlab = "t", ylab = "",
     main = "Cash Flow Timeline",
     xaxt = "n", yaxt = "n")


axis(1, at = time_periods)


for (i in 1:length(cash_flows)) {
  lines(c(i-1, i-1), c(0, cash_flows[i]))
  
  
  if (i <= 4 || i >= 11) {
    
    label <- round(cash_flows[i], 2)
  } else if (i == 7) {
    
    label <- "..."
  } else {
   
    next
  }
  
  text(i-1, cash_flows[i], labels = label, 
       pos = ifelse(cash_flows[i] > 0, 3, 1),
       offset = 0.5)
}

# Add horizontal line
abline(h = 0)
```

The net PV is ~0, meaning the PV of the repayments equals the PV of the loans, so the yearly payment is `r round(yearly_payment, 4)`.

#### Simulation for Yearly Payments Analysis

To assess the variability in yearly payments, we can simulate different interest rate scenarios.

```{r}
# Simulation for varying interest rates
set.seed(123)
n_simulations <- 1000
simulated_payments <- replicate(n_simulations, {
  interest_sim <- c(rep(runif(1, 0.035, 0.045), 3), rep(runif(1, 0.04, 0.05), 3), rep(runif(1, 0.055, 0.065), 5))
  yearly_discount_factors_sim <- (1 + interest_sim) ^ (-1)
  discount_factors_sim <- cumprod(c(1, yearly_discount_factors_sim))
  
  PV_loan_sim <- sum(cash_flow * discount_factors_sim[1:2])
  repayment_years_sim <- 2:11
  discount_factors_repayments_sim <- discount_factors_sim[repayment_years_sim + 1]
  yearly_payment_sim <- PV_loan_sim / sum(discount_factors_repayments_sim)
  
  return(yearly_payment_sim)
})

# Calculate the mean and standard deviation of simulated yearly payments
mean_payment <- mean(simulated_payments)
sd_payment <- sd(simulated_payments)
cat("Mean Yearly Payment:", mean_payment, "\n")
cat("Standard Deviation of Yearly Payment:", sd_payment, "\n")

# Plot the distribution of simulated yearly payments
hist(simulated_payments, breaks = 30, main = "Distribution of Simulated Yearly Payments", xlab = "Yearly Payment")
abline(v = mean_payment, col = "red", lwd = 2)


```

------------------------------------------------------------------------------------

## Yearly to Monthly Interest Rate

You plan to take out a 100,000 loan with a yearly interest rate of 2.88%. You want to pay back this loan with fixed monthly payments over the next 15 years. Find how much these monthly payments will be.

```{r}

#number of payments
number_payments <- 12*15

#yearly interest rate
i <- 0.0288

#monthly interest rate

monthly_interest <- (1+i)^(1/12)-1


discount_factors <- (1 + monthly_interest) ^ - (1:number_payments)


payments <- rep(1, number_payments)

# Calculate the monthly loan payment K
K <- 100000 / sum(payments * discount_factors)
cat("Monthly loan payments =", K)
```
These monthly payments will be `r round(K, 4)`.

#### Monthly Payments Under Interest-Rate Uncertainty

Once the loan is signed the payment is fixed, but before you lock in a rate you don't know exactly what you'll be offered. Assuming the rate you're offered is normally distributed around 2.88% with a standard deviation of 0.25% (an illustrative assumption), we can simulate the range of monthly payments.

```{r}
# Simulated monthly payments for uncertain interest rates
set.seed(123)
n_sim <- 10000
sim_payments <- replicate(n_sim, {
  i_sim <- rnorm(1, mean = 0.0288, sd = 0.0025)
  monthly_sim <- (1 + i_sim)^(1/12) - 1
  K_sim <- 100000 / sum((1 + monthly_sim) ^ - (1:number_payments))
  return(K_sim)
})

# Middle 95% of simulated payments
ci_payments <- quantile(sim_payments, c(0.025, 0.975))
cat("Middle 95% of monthly payments:", ci_payments)


```


#### Could the payment exceed $700?

```{r}
cat("Probability the monthly payment exceeds $700:", mean(sim_payments > 700), "\n")
rate_for_700 <- uniroot(function(r) 100000 / sum((1 + ((1 + r)^(1/12) - 1)) ^ -(1:number_payments)) - 700,
                        c(0, 0.2))$root
cat("Yearly rate at which the payment reaches $700:", round(100 * rate_for_700, 3), "%")

```

---------------------------------------------------------------------------------

# Life Tables

<details>
  <summary>What are Life Tables?</summary>
  <p style="font-weight: bold; color: blue;">
  Life tables are statistical tools used in demography and actuarial science to analyze mortality rates and life expectancy within a population. They provide a detailed breakdown of how long people are expected to live, based on age-specific mortality rates. Life tables tell us about the probability of surviving to different ages, expected years of life remaining at various ages, and are crucial for understanding population dynamics, healthcare planning, and insurance risk assessment.In data analysis beyond actuarial science, life tables serve as a model for understanding survival or retention rates in various contexts. For instance, in customer analytics, they can help predict how long customers are likely to remain active or how likely they are to churn based on historical data. Similarly, in medical research, life tables can be adapted to study survival rates after treatments or interventions. Essentially, life tables provide a structured way to analyze and predict durations or outcomes based on observed data, making them versatile tools across different fields of data analysis.
   </p>
</details>

**Import the life tables from [mortality.org](https://mortality.org/). We are going to import the dataset of Canada from 1921 - 2021.**
```{r}
# Loaded in the setup chunk (see load_life_table())
life_table <- life_table_all
head(life_table)
```


**Setting up variables** 
```{r}
year <- life_table$Year
age <- life_table$Age
qx <- life_table$qx
mx <- life_table$mx
ax <- life_table$ax
lx <- life_table$lx
dx <- life_table$dx
Lx <- life_table$Lx
Tx <- life_table$Tx
ex <- life_table$ex
px <- 1 - qx
```


| **Variable** | **Description** |
|--------------|-----------------|
| Year         | Year or range of years (for period and cohort data) |
| Age          | Age group, representing an n-year interval from exact age x to just before x + n, where n = 1, 4, 5, or infinity |
| m(x)         | Central death rate between ages x and x + n |
| q(x)         | Probability of death between ages x and x + n |
| a(x)         | Average length of survival between ages x and x + n for persons dying in the interval |
| l(x)         | Number of survivors at exact age x, assuming l(0) = 100,000 |
| d(x)         | Number of deaths between ages x and x + n |
| L(x)         | Number of person-years lived between ages x and x + n |
| T(x)         | Number of person-years remaining after exact age x |
| e(x)         | Life expectancy at exact age x (in years) |


------------------------------------------------------------------------------------

## Mortality Rates

**Plotting the mortality rates from 1921 to 2021 for all ages**

```{r}
# Setting up the plot
par(xpd = TRUE, mar = c(5, 4, 4, 6) + 0.1) 

convert_age <- function(x) {
  x[x == "110+"] <- "110"  # Here age is character in our dataset, need to change to numeric so that we don't run into issues with character 
  suppressWarnings({
    as.numeric(x)
  })
}
age <- convert_age(age)


years <- unique(year)
color_palette <- colorRampPalette(c("blue", "green", "yellow", "red"))(length(years))


plot2 <- plot(NULL, 
              xlim = range(age, na.rm = TRUE),  
              ylim = range(qx[qx > 0], na.rm = TRUE),   
              log = "y",  # log scale: qx spans 0.0001 to 1
              main = "Mortality rates (Canada, both sexes, 1921-2021)",
              xlab = "Age x",
              ylab = expression(paste("Mortality rate ", q[x], " (log scale)")),
              type = "l"
)


suppressWarnings({
  for (i in seq_along(years)) {
    year_data <- subset(life_table, Year == years[i])
    lines(year_data$Age, year_data$qx, col = color_palette[i], type = "l")
  }
})


legend("topright", inset = c(-0.2,0),
       legend = c("1921", "1955", "1988", "2021"),
       col = color_palette[c(1, which.min(abs(years - 1955)), 
                             which.min(abs(years - 1988)), length(years))],
       lty = 1, 
       title = "Year")

```

On a log scale the improvement is clear at every age: mortality in recent years is far lower than in the earliest years, with the biggest falls in infancy and childhood. 

#### Hypothesis Testing on Mortality Rates

We can perform a two-proportion z-test to check whether the mortality rate at age 50 changed between 1921 and 2021. One caveat: a life table's lx counts survivors out of a hypothetical 100,000 births, not the real population, so using it as the sample size makes this an illustration. A proper test would use the actual death counts and exposures (the HMD `Deaths_1x1` and `Exposures_1x1` files).

```{r}
# Hypothesis testing for mortality rates at age 50
row_1921 <- subset(life_table, Year == 1921 & Age == 50)
row_2021 <- subset(life_table, Year == 2021 & Age == 50)
mort_1921 <- row_1921$qx
mort_2021 <- row_2021$qx
# two-proportion z-test using each year's survivors at age 50 as the sample size
se <- sqrt(mort_1921 * (1 - mort_1921) / row_1921$lx + mort_2021 * (1 - mort_2021) / row_2021$lx)
t_stat <- (mort_2021 - mort_1921) / se
p_value <- 2 * pnorm(-abs(t_stat))
cat("q50 in 1921:", mort_1921, "  q50 in 2021:", mort_2021, "\n")
cat("z-statistic:", t_stat, "\n")
cat("P-value:", p_value, "\n")

if (p_value < 0.05) {
  cat("We reject the null hypothesis that the mortality rate at age 50 has not changed between 1921 and 2021\n")
} else {
  cat("We fail to reject the null hypothesis that the mortality rate at age 50 has not changed between 1921 and 2021\n")
}
```

------------------------------------------------------------------------------------

## Life Expectancy

**Plotting the life expectancy in Canada (both sexes combined) from the years 1921 - 2021**

```{r}

par(xpd = TRUE, mar = c(5, 4, 4, 6) + 0.1) 


years <- unique(year)
color_palette <- colorRampPalette(c("blue", "green", "yellow", "red"))(length(years))

plot3 <- suppressWarnings({
  plot(NULL, 
       xlim = range(age, na.rm = TRUE),  
       ylim = range(ex, na.rm = TRUE),   
       main = "Life Expectancy (Canada, both sexes, 1921-2021)",
       xlab = "Age x",
       ylab = expression(paste("Life Expectancy ", e[x])),
       type = "l"
  )
})


suppressWarnings({
  for (i in seq_along(years)) {
    year_data <- subset(life_table, Year == years[i])
    lines(year_data$Age, year_data$ex, col = color_palette[i], type = "l")
  }
})


legend("topright", inset = c(-0.2, 0),
       legend = c("1921", "1955", "1988", "2021"),
       col = color_palette[c(1, which.min(abs(years - 1955)), 
                             which.min(abs(years - 1988)), length(years))],
       lty = 1, 
       title = "Year")



```

As we can see, the life expectancy is much higher from the years ~ 2021 as opposed to the years from ~1921 or even ~1955. 
Note for the early years, the life expectancy starts extremely low at age 0 and after a couple years, has a sharp increase. 
This can be explained mainly because it was much more difficult for an infant to survive the initial years. In later years, thanks
to modern medicine, the survival rate of infants have risen much higher.


#### Life Expectancy at Birth Over Time

Each year has exactly one life expectancy at birth, calculated from that year's mortality rates, so these are not repeated samples of one quantity. Averaging them, or bootstrapping them as an earlier version did, mixes a century of very different conditions into a single number that doesn't describe any real population. It is more informative to look at the trend.

```{r}
e0 <- subset(life_table, Age == 0)
plot(e0$Year, e0$ex, type = "l", lwd = 2, col = "darkgreen",
     xlab = "Year", ylab = expression(e[0]), main = "Life expectancy at birth (Canada, both sexes)")
trend <- lm(ex ~ Year, data = e0)
abline(trend, lty = 2)
cat("Life expectancy at birth: 1921 =", e0$ex[e0$Year == 1921], "  2021 =", e0$ex[e0$Year == 2021], "\n")
cat("Average gain per decade (linear trend):", round(10 * coef(trend)[["Year"]], 2), "years\n")
```

Life expectancy at birth rose by about `r round(10 * coef(trend)[["Year"]], 1)` years per decade over the century. The dip at the end is 2020-2021, the COVID-19 pandemic.


Let us focus on the year 2021.

**2021 Life Table**
```{r}
life_table_2021 <- subset(life_table_all, Year == 2021)  #Getting only data from the year 2021
year <- life_table_2021$Year
age <- life_table_2021$Age
qx <- life_table_2021$qx
mx <- life_table_2021$mx
ax <- life_table_2021$ax
lx <- life_table_2021$lx
dx <- life_table_2021$dx
Lx <- life_table_2021$Lx
Tx <- life_table_2021$Tx
ex <- life_table_2021$ex
px <- 1 - qx
```


### Probability an individual of age X dies before age Y, future life of age X

**Finding the probability that an 23 year old dies before turning 24**

```{r}

cat("qx =",qx[23 + 1])

```

As we can see, it is a very low probability: about `r round(100 * qx[23 + 1], 3)`%.

**The expected future lifetime of an 23 year old**
```{r}
cat("ex =",ex[23 + 1])

```
Based on 2021 mortality rates, an average 23 year old in Canada is expected to live `r ex[23 + 1]` more years.

### Plotting mortality rates

**Plot the mortality rates in the year 2021**

```{r}
suppressWarnings({
  plot(age, qx, 
       main = "Mortality rates (Canada, 2021)", 
       xlab = "Age x", 
       ylab = expression(paste("Mortality rate ", q[x])), 
       type = "l")
})

```


**Plot the logarithm of the mortality rates in the year 2021**
```{r}

suppressWarnings({
  plot(age, log(qx), 
       main = "Log mortality rates (Canada, 2021)", 
       xlab = "Age x", 
       ylab = expression(paste("Log mortality rate ", log(q[x]))), 
       type = "l")
})

```

------------------------------------------------------------------------------------

**Plot the mortality rates of females vs males in the year 2021**

We need to import the male and female datasets here

```{r, eval = has_sex_tables}
# Male and female tables are loaded in the setup chunk from data/mltper_1x1.txt
# and data/fltper_1x1.txt (HMD, Canada). Sections using them are skipped if absent.
male_2021 <- subset(life_table_m_all, Year == 2021)
age_m <- male_2021$Age
qx_m <- male_2021$qx
lx_m <- male_2021$lx
ex_m <- male_2021$ex

female_2021 <- subset(life_table_f_all, Year == 2021)
age_f <- female_2021$Age
qx_f <- female_2021$qx
lx_f <- female_2021$lx
ex_f <- female_2021$ex
```

```{r, echo = FALSE, results = "asis", eval = !has_sex_tables}
cat("> *The male and female life tables (`data/mltper_1x1.txt`, `data/fltper_1x1.txt`) are not included in this folder, so the male vs female comparisons below are skipped. Download them from mortality.org (Canada, period 1x1) to enable them.*")
```

**Plot male vs female mortality rates in the year 2021**

```{r, eval = has_sex_tables}
suppressWarnings({
  plot(age_m, qx_m, 
       main = "Mortality rates (Canada, 2021)", 
       xlab = "Age x", 
       ylab = expression(paste("Mortality rate ", q[x])), 
       type = "l",
       col = "blue",
  )  
  
  # Add the female line to the same plot
  lines(age_f, qx_f, col = "red")
  
  legend("bottomright", 
         legend = c("Male", "Female"),
         col = c("blue", "red"),
         lty = 1,
         cex = 0.8)
})


```

```{r, eval = has_sex_tables}


suppressWarnings({
  plot(age_m, qx_m, 
       main = "Mortality rates (Canadian Males and Females, 2021)", 
       xlab = "Age x", 
       ylab = expression(paste("Mortality rate ", q[x])), 
       type = "l",
       col = "blue",
       log = "y",  
       ylim = range(c(qx_m, qx_f)) 
  )
  
  lines(age_f, qx_f, col = "red")
  
  grid()
  
  points(age_m, qx_m, pch = 20, cex = 0.5, col = "blue")
  points(age_f, qx_f, pch = 20, cex = 0.5, col = "red")
  
  legend("bottomright", 
         legend = c("Males", "Females"),
         col = c("blue", "red"),
         lty = 1,
         pch = 20,
         cex = 0.8)
})

```

As we can see, females on average for every age have a lower mortality rate than males. The gap only begins to narrow at the later ages, where it eventually becomes near equal around the age of 100. While there could be multiple reasons for this disparity, a study from [Harvard](https://www.health.harvard.edu/blog/why-men-often-die-earlier-than-women-201602199137) states that one of the reasons could be because men have lower estrogen levels, hence they are more likely to die of heart disease

### Probability for invidivual of age x to reach age y

**Find the probability for a 0 year old female and male in Canada to reach the age of 100 in 2021**

```{r, eval = has_sex_tables}
male_prob <- lx_m[100 + 1] / lx_m[0 + 1]
print(paste("Male probability:", male_prob))


female_prob <- lx_f[100 + 1] / lx_f[0 + 1]
print(paste("Female probability:", female_prob))
```

------------------------------------------------------------------------------------

**Find the probability for a 0 year old in Canada to reach the age of 100 in 2021**

```{r}
canada_prob <- lx[100 + 1] / lx[0 + 1]
print(paste("M/F probability:", canada_prob))
```

If we compare this with an 18 year old individual

```{r}
canada_prob_2 <- lx[100 + 1] / lx[18 + 1]
print(paste("M/F age 18 probability:", canada_prob_2))
```
Vs a 40 year old individual

```{r}
canada_prob_5 <- lx[100 + 1] / lx[40 + 1]
print(paste("M/F age 40 probability:", canada_prob_5))
```



And again with a 75 year old individual

```{r}
canada_prob_3 <- lx[100 + 1] / lx[75 + 1]
print(paste("M/F age 75 probability:", canada_prob_3))
```
Finally, compare with a 98 year old individual

```{r}
canada_prob_4 <- lx[100 + 1] / lx[98 + 1]
print(paste("M/F age 98 probability:", canada_prob_4))
```

```{r}

ages <- c(0, 18, 40, 75, 98)
probs <- c(
  lx[100 + 1] / lx[0 + 1],
  lx[100 + 1] / lx[18 + 1],
  lx[100 + 1] / lx[40 + 1],
  lx[100 + 1] / lx[75 + 1],
  lx[100 + 1] / lx[98 + 1]
)


prob_data <- data.frame(Age = c("0", "18", "40", "75", "98"), Probability = probs)


barplot(prob_data$Probability, 
        names.arg = prob_data$Age,
        main = "Probability of Reaching Age 100 from Different Starting Ages",
        xlab = "Starting Age",
        ylab = "Probability",
        col = "skyblue",
        ylim = c(0, max(probs) * 1.1))  


text(x = 1:5, 
     y = probs,
     labels = round(probs, 5),
     pos = 3,
     cex = 0.8)

grid()
# Just adding a grid for some better readability

```


**Plot the survival probabilties for (18) up to age 100**
```{r}
k <- 0:82
plot(k, lx[18 + k + 1] / lx[18 + 1], 
    pch = 20, 
    xlab = "k", 
    ylab = expression(paste(""[k], "p"[18])), 
    main = "Survival probabilities for age 18 up to 100")
```

Compare this with a 50 year old

```{r}
k <- 0:50
plot(k, lx[50 + k + 1] / lx[50 + 1], 
    pch = 20, 
    xlab = "k", 
    ylab = expression(paste(""[k], "p"[50])), 
    main = "Survival probabilities for age 50 up to 100")
```

**Plot the number of deaths**
```{r}

suppressWarnings({
  plot(age, dx, 
       type = "h", 
       pch = 20, 
       xlab = "Age x", 
       ylab = expression("d"[x]),
       main = "Number of deaths (Canada, M/F, 2021)")
  
  # one simulated set of deaths: each of the lx survivors dies with probability qx
  set.seed(123)
  sims <- rbinom(n = length(lx), size = lx, prob = qx)
  points(age, sims, 
         pch = 4, 
         col = "red")
  legend("topleft", legend = c("Life table deaths", "Simulated (binomial)"),
         col = c("black", "red"), lty = c(1, NA), pch = c(NA, 4))
})


```

### Probability that individual of age x survives y more years

**Calculate the probability that (25) survives 5 more years**

<details>
  <summary style="font-weight: bold; color: blue;">Explanation</summary>
  <p>
    <strong>px Calculation:</strong><br>
    px typically represents the survival probability (1 - qx) at each age x. In this context, it's assumed to be a sequence or vector containing survival probabilities for each age from 25 onwards.<br><br>
    
    <strong>Subset of px:</strong><br>
    px[(25 + 1):(29 + 1)] selects p25, p26, p27, p28 and p29 (R vectors start at 1, so age x is stored at position x + 1). These are the five one-year survival probabilities from age 25 to age 30.<br><br>
    <strong>Product Calculation:</strong><br>
    prod(px[(25 + 1):(29 + 1)]) multiplies them: surviving five years in a row means surviving each single year, so 5p25 = p25 × p26 × p27 × p28 × p29.<br><br>
    
    <strong>Output:</strong><br>
    cat("Prob that a 25 year old survives for 5 more years =", value1) prints the result to the console in a human-readable format. It displays the probability that a 25-year-old is still alive at age 30.
  </p>
</details>



```{r}
value1 <- prod(px[(25 + 1):(29 + 1)])
cat("Prob that a 25 year old survives for 5 more years =", value1)
```

*How to interpret these results*: 

Each one-year survival probability at these ages is very close to 1, so even five years in a row gives a probability of about `r round(100 * value1, 2)`%. Survival probabilities fall as age increases: in 2021, p30 = `r px[30 + 1]` (`r round(100 * px[30 + 1], 2)`% of 30-year-olds reach 31), p50 = `r px[50 + 1]`, p70 = `r px[70 + 1]` and p90 = `r round(px[90 + 1], 4)`. The decline is slow through middle age and much steeper after about 70, reflecting the increasing risk of death at older ages.

**Compute the survival probabilities of (25) until the age of 100**

<details>
  <summary style="font-weight: bold; color: blue;">Explanation</summary>
  <p>
    <strong>Survival Probability (px):</strong><br>
    px is a vector where each element represents the probability of survival (1 - qx) at a specific age x. It starts from age 25 and continues for subsequent ages.<br><br>

    <strong>Subset of Survival Probabilities (px[(25 + 1):(99 + 1)]):</strong><br>
    This selects p25 through p99 (positions 26 to 100), the one-year survival probabilities needed to get from age 25 to age 100.<br><br>

    <strong>Cumulative Product Calculation (cumprod(px[(25 + 1):(99 + 1)])):</strong><br>
    cumprod calculates the cumulative product of the selected survival probabilities. The k-th value is kp25, the probability that a 25-year-old survives k more years (to age 25 + k). The last value is the probability of reaching 100.
  </p>
</details>




```{r}
kpx <- cumprod(px[(25 + 1):(99 + 1)])
cat("Survival prob's of a 25 year old until 100 =", kpx)
```

*How to interpret these results*:

The output kpx gives the probability of a 25-year-old reaching each age from 26 to 100. For example, if kpx[1] is 0.9992, it means there's a 99.92% chance that a 25-year-old individual will survive until age 26. Similarly, kpx[2] (0.9984006) represents the probability of surviving from age 25 to age 27, and so on.


**Extract the probability that (25) survives until the age of 100**

```{r}
value3 <- kpx[length(kpx)]
cat("Prob that a 25 year old survives until 100:", value3)
```


**Plot the probabilties for (25) to reach the age of 26, 27, ..., 100**

```{r}
plot(1:length(kpx), kpx,
    pch = 20,
    xlab = "k",
    ylab = expression(paste(""[k], "p"[25])),
    main = "Survival probabilities for (25)")
```

## Survival Probabilities and Curtate Expected Future Lifetime

<details>
  <summary>What is Curtate Expected Future Lifetime?</summary>
  <p style="font-weight: bold; color: blue;">
  Curtate expected future lifetime is a statistical measure used in actuarial science and life insurance. It represents the expected number of whole years a person, starting at a given age, is expected to live into the future. This measure is calculated by considering the probability of surviving each year from the current age until the individual's death. Curtate expected future lifetime provides valuable insights into life expectancy and is crucial for determining insurance premiums and retirement planning.
   </p>
</details>


**Compute the survival probabilities of (25)**

```{r}
kpx <- c(1, cumprod(px[(25 + 1):(length(px) - 1)]))
cat("Survival prob's of a 25 year old:", kpx)
```

**Compute the deferred mortality probabilities of (25)**

<details>
  <summary style="font-weight: bold; color: blue;">Explanation</summary>
  <p>
    <strong>Deferred Mortality Probabilities of a 25-Year-Old:</strong><br>
    `kqx[k + 1]` is the deferred mortality probability k|q25: the probability that a 25-year-old survives exactly k more years and then dies in the following year, i.e. dies between ages 25 + k and 25 + k + 1.<br><br>
    <strong>Calculation:</strong><br>
    It is the probability of surviving k years, kp25, times the probability of dying in the next year at the age then reached, q(25 + k): k|q25 = kp25 × q(25 + k).<br><br>
    <strong>Check:</strong><br>
    Everyone dies at some age, so these probabilities add up to 1 (q110 = 1 closes the table). In the chart below, the peak shows the most likely age at death for someone who is 25 now.<br><br>
  </p>
</details>


```{r}
kqx <- kpx * qx[(25 + 1):length(qx)]
cat("Deferred mortality prob's of a 25 year old:", kqx, "\n")
cat("Sum (should be 1):", sum(kqx), "\n")
cat("Most likely age at death:", 25 + which.max(kqx) - 1)
```

<details>
  <summary style="font-weight: bold; color: blue;">Why Use Deferred Mortality Probabilities?</summary>
  <p>
    <strong>Risk Assessment:</strong> Deferred mortality probabilities provide a more accurate assessment of mortality risk over time. They give the full distribution of the age at death, not just its average.<br><br>

    <strong>Product Design:</strong> Insurance companies and pension funds use deferred mortality probabilities to design products such as life insurance policies and annuities. These products rely on accurate estimates of how mortality risk changes with age to determine premiums and payouts.<br><br>

    <strong>Financial Planning:</strong> Individuals and institutions use deferred mortality probabilities for retirement planning and financial forecasting. They help in estimating life expectancies and determining the financial resources needed for retirement based on expected future lifetimes.<br><br>

    <strong>Actuarial Calculations:</strong> Actuaries use deferred mortality probabilities in complex calculations such as pension fund valuations, reserve calculations, and risk assessments. These calculations ensure that financial institutions have adequate reserves to cover future obligations.<br><br>

    <strong>Long-Term Commitments:</strong> Many financial products, such as pensions and annuities, involve long-term commitments spanning several decades. Deferred mortality probabilities provide insights into how mortality risk impacts these long-term financial commitments.<br><br>

    Overall, deferred mortality probabilities are essential tools in understanding and managing mortality risk in various financial contexts. They give the timing of deaths, and therefore of insurance payouts, informing decisions that impact both individuals and financial institutions over the long term.
  </p>
</details>

**Plot the deferred mortality probabilities of (25)**

```{r}
plot(0:(length(kqx) - 1), kqx,
    pch = 20,
    xlab = "k",
    ylab = expression(paste(""['k|'], "q"[25])),
    main = "Deferred mortality probabilities of (25)")
```


**Survival probabilities and curtate expected future lifetime of (0)**

```{r}
kp0 <- cumprod(px)
cat("Survival prob's of a 0 year old", kp0)
sum_kp0 <- sum(kp0)
cat("Curtate expected future lifetime of a 0 year old",sum_kp0 )
```

**Survival probabilities and curtate expected future lifetime of a 25 year old**

```{r}
kp25 <- cumprod(px[(25 + 1):length(px)])
cat("Survival prob's of a 25 year old", kp25)
sum_kp25 <- sum(kp25)
cat("Curtate expected future lifetime of a 25 year old",sum_kp25 )

```

**Complete expected future lifetime of (0) and (25)**

```{r}
efl <- ex[c(0 + 1, 25 + 1)]
cat("Expected future lifetime of a 0 and 25 year old", efl)
```

### Function to compute the curtate expected future lifetime for a given age and life table

<details>
  <summary style="font-weight: bold; color: blue;">Explanation: Computing Curtate Expected Future Lifetime</summary>
  <p>
    <strong>Function Purpose:</strong> The function `curtate_future_lifetime` calculates the curtate expected future lifetime for a given starting age using a life table dataset.<br><br>

    <strong>Inputs:</strong> 
    
    - `age2`: Represents the starting age for which we want to compute the curtate expected future lifetime.
    - `life_table`: Contains mortality data including qx (probability of dying between ages x and x+1) needed for calculations.<br><br>

    <strong>Steps:</strong>
    
    1. <strong>qx Calculation:</strong> Extracts qx values from the `life_table`. qx represents the probability of dying between ages x and x+1.<br>
    2. <strong>px Calculation:</strong> Computes px as 1 - qx, which represents the probability of survival between ages x and x+1.<br>
    3. <strong>Subset and Cumulative Product:</strong> Selects and computes the cumulative product of px values starting from `age2 + 1` to the end of the dataset. This sequence represents the joint probability of surviving each subsequent year after `age2`.<br>
    4. <strong>Summation:</strong> Sums up the cumulative product (`kpx`) to obtain the curtate expected future lifetime, which represents the expected number of full years an individual aged `age2` is expected to live from that age.<br><br>

    <strong>Output:</strong> The function returns the computed curtate expected future lifetime as a numeric value.<br><br>

    This function is crucial in actuarial science and insurance for estimating life expectancies and making financial projections based on mortality data.
  </p>
</details>


```{r, echo=TRUE}
suppressWarnings({
  curtate_future_lifetime <- function(age2, life_table) {
    qx <- life_table$qx
    px <- 1 - qx
    kpx <- cumprod(px[(age2 + 1):length(px)])
    sum(kpx)
  }
})

```


**Curtate future lifetimes for all ages**

```{r}
age2 <- life_table_2021$Age
valid_ages <- age2[!is.na(age2)]
future_lifetimes <- sapply(valid_ages, function(age) curtate_future_lifetime(age, life_table_2021))
cat("Future lifetimes for all ages:", future_lifetimes)
```

*How to interpret*:

These are the expected number of years an individual, starting from each age, is expected to live based on the curtate expected future lifetime calculation.Each value corresponds to the expected future lifetime for individuals starting from different ages.

For example, the first value, `r round(future_lifetimes[1], 2)`, is for a newborn: about `r round(future_lifetimes[1], 2)` complete years of future life. For a 25-year-old (the 26th value) it is `r round(future_lifetimes[26], 2)`. Curtate values are about half a year lower than the complete life expectancy ex, because they count only whole years lived.
The values decrease as the starting age increases, reflecting the decreasing life expectancy as individuals age.
These estimates are based on mortality data and assume that future mortality rates remain consistent with those observed in the life table used for calculation.

```{r}
# Future lifetime by age
plot(age2[!is.na(age2)], future_lifetimes, type = 'l', lwd = 2, col = "green", xlab = "Age x", ylab = "Future lifetime", main = "Future lifetime by age")
```


**Plot the logarithm of the mortality rates for (18) by year (both sexes)**

```{r}

suppressWarnings({
  life_table <- life_table_all

  # Subset the life table for age 18
  subset_18 <- subset(life_table, Age == 18)

  
  if (nrow(subset_18) > 0) {
    with(subset_18,
         plot(Year, log(qx),
              type = "l", main = "Log mortality rates (Canada, both sexes, age 18)",
              xlab = "Year t", ylab = expression(paste("Log mortality rate ", log(q[18])))))
  } else {
    print("No data available for age 18")
  }
})


```



**Plot the logarithm of the mortality rates in the year 1950 by age (both sexes)**

```{r}
# Subset the life table for the year 1950
subset_1950 <- subset(life_table, Year == 1950)


if (nrow(subset_1950) > 0) {
  with(subset_1950,
       plot(Age, log(qx),
            type = "l", main = "Log mortality rates (Canada, both sexes, 1950)",
            xlab = "Age x", ylab = expression(paste("Log mortality rate ", log(q[x])))))
} else {
  print("No data available for the year 1950")
}
```

**Constructing and printing the cohort life table for birth year 1981**

A cohort table follows one birth year through time (age 0 in 1981, age 1 in 1982, ... age 40 in 2021), instead of taking all ages from a single calendar year as a period table does. Because the data end in 2021, the 1981 cohort is only observed up to age 40.

```{r}
life_table_1981 <- subset(life_table, Year - Age == 1981)
life_table_1981
```

**1981 cohort one-year survival probabilities**

```{r}
px <- 1 - life_table_1981$qx
cat("One year survival probabilities for 1981 birth year = ", px)
```

**1981 cohort survival probability that (18) survives 5 more years**

```{r}
prodp <- with(life_table_1981, prod(1 - qx[Age %in% 18:22]))
cat("Survival probability that an 18 year old survives 5 more years for 1981 birth year =", prodp)
```

**1903 cohort survival probability that (18) survives 5 more years** (the data start in 1921, so 1903 is the earliest birth year observed at age 18)

```{r}
prodp2 <- with(subset(life_table, Year - Age == 1903 & Age %in% 18:22), prod(1 - qx))
cat("Survival probability that an 18 year old survives 5 more years for 1903 birth year =", prodp2)

```


**PV of guaranteed payment of 10,000 in 5 years**

```{r}
PV <- 10000 * (1 + 0.02) ^ -5
cat("PV =",PV)
```

**5 year survival probabilities of (20)**

```{r}
# Back to the 2021 period life table for the actuarial calculations
qx <- life_table_2021$qx
px <- 1 - qx
kpx <- prod(px[(20 + 1):(24 + 1)])
cat("kpx =", kpx)
```

**EPV of pure endowment of 10,000 in 5 years for (20)**

```{r}
epv <- PV * kpx
cat("Expected Present value =", epv)
```

# Life Annuities

**PV of guaranteed payments of 10,000 in 5, 10 and 30 years**

```{r}
PV <- 10000 * (1 + 0.02) ^ - c(5, 10, 30)
cat("Present value =",PV)
```

**Survival probabilities of (20)**

```{r}
kpx <- cumprod(px[(20 + 1):length(px)])
cat("kpx =", kpx)
```

**EPV of pure endowments of 10,000 in 5, 10 and 30 years for (20)**

```{r}
epv2 <- PV * kpx[c(5, 10, 30)]
cat("Expected Present Value =", epv2)
```

## Whole Life Annuity Due

A type of annuity where payments are made at the beginning of each period, starting immediately upon purchase and continuing for the lifetime of the annuitant. Payments start immediately and occur at the beginning of each payment period (e.g., monthly, annually)

**Calculating the whole life annuity due of a 35 year old**

```{r}
kpx <-
  c(1, cumprod(px[(35+1):length(px)]))
discount_factors <- (1+0.03)^-(0:(length(kpx)-1))
benefits <- rep(1,length(kpx))
wld <- sum(benefits*discount_factors*kpx)
cat("Whole life annuity due =", wld)

```

## Whole Life Immediate Annuity

A type of annuity where payments are made at the end of each period, starting immediately upon purchase and continuing for the lifetime of the annuitant. Payments start immediately but occur at the end of each payment period (e.g., monthly, annually).

**Calculating the whole life immediate annuity of a 35 year old**

```{r}
kpx <- cumprod(px[(35+1):length(px)])
discount_factors <- (1+0.03)^-(1:(length(kpx)))
benefits <- rep(1,length(kpx))
wid <- sum(benefits*discount_factors*kpx)
cat("Whole life immediate annuity =", wid)
```

## Expected Present Value 

#### Function to compute the EPV of a whole life annuity due for a given age, interest rate i and life table

```{r}
life_annuity_due <- function(age, i, lt) {
  q <- lt$qx[(age + 1):nrow(lt)]              # q_x, q_(x+1), ..., q_110 (= 1)
  kpx <- c(1, cumprod(1 - q))[seq_along(q)]   # k p_x for k = 0, 1, 2, ...
  sum((1 + i) ^ -(seq_along(q) - 1) * kpx)    # payments at the start of each year
}
```

## Shiny App for Life Annuity Calculator


<iframe src="https://camleo.shinyapps.io/life_annuity_app" width="100%" height="300"></iframe>



**EPV of a whole life annuity due for (20) at interest rate 2% using the 2021 life table**

```{r}
epv3 <- life_annuity_due(20, 0.02, life_table_2021)
cat("EPV of whole life annuity due =", epv3)
```

**EPV of a whole life annuity due for (20) at interest rate 5% and for (65) at interest rate 2% using the 2021 life table**

```{r}
epv4 <- life_annuity_due(20, 0.05, life_table_2021)
cat("EPV of life annuity due for 20 year old at 5% interest rate =", epv4)
epv5 <- life_annuity_due(65, 0.02, life_table_2021)
cat("EPV of life annuity due for 65 year old at 2% interest rate =", epv5)

```


#### Function to compute the EPV of a whole life immediate annuity for a given age, interest rate i and life table

```{r}
life_immediate_annuity <- function(age, i, lt) {
  q <- lt$qx[(age + 1):nrow(lt)]
  kpx <- cumprod(1 - q)                       # k p_x for k = 1, 2, ...
  sum((1 + i) ^ -seq_along(q) * kpx)          # payments at the end of each year
}
```

**EPV of a whole life immediate annuity for (20) at interest rate 2% using the 2021 life table**

```{r}
epv6 <- life_immediate_annuity(20, 0.02, life_table_2021)
cat("EPV of life annuity immediate for a 20 year old using 2% interest rate =", epv6, "\n")
# check: the immediate annuity is the annuity due without the first payment
cat("Annuity due - 1 =", epv3 - 1)
```



#### Function to compute the EPV of a temporary life annuity due for a given age, period of n years, interest rate i and life table

```{r}
temporary_life_annuity_due <- function(age, n, i, lt) {
  q <- lt$qx[(age + 1):nrow(lt)]
  n <- min(n, length(q))                      # can't pay beyond the end of the table
  kpx <- c(1, cumprod(1 - q))[1:n]            # k p_x for k = 0, ..., n - 1
  sum((1 + i) ^ -(0:(n - 1)) * kpx)
}
```

## Shiny app for Temporary Life Annuity Calculator

```{r, eval = FALSE}
library(shiny)

ui <- fluidPage(
  titlePanel("Temporary Life Annuity Calculator"),
  sidebarLayout(
    sidebarPanel(
      numericInput("age", "Age:", 20, min = 0, max = 110),
      numericInput("term", "Term (years):", 10, min = 1, max = 50),
      numericInput("interest", "Interest Rate:", 0.02, min = 0, max = 0.2, step = 0.001)
    ),
    mainPanel(
      h3("Expected Present Value:"),
      textOutput("epv")
    )
  )
)

server <- function(input, output) {
  output$epv <- renderText({
    epv <- temporary_life_annuity_due(input$age, input$term, input$interest, life_table_2021)
    paste0("$", round(epv, 2))
  })
}

shinyApp(ui = ui, server = server)
```

*The live Shiny app can't run on this static GitHub Pages site. Open `index.Rmd` in RStudio, run the earlier chunks, then run this chunk to launch it. Sample results:*

```{r, echo = FALSE}
grid <- expand.grid(Age = c(20, 40, 65), Term = c(10, 20), Interest = c(0.02, 0.05))
grid$EPV <- round(mapply(temporary_life_annuity_due, grid$Age, grid$Term, grid$Interest,
                         MoreArgs = list(lt = life_table_2021)), 4)
knitr::kable(grid, caption = "EPV of a temporary life annuity due of 1 per year (2021 Canada table)")
```




**EPV of a temporary life annuity due for (20) over 10 years at interest rate 2% using the 2021 life table**

```{r}
epvt <- temporary_life_annuity_due(20, 10, 0.02, life_table_2021)
cat("EPV of temporary life annuity due for 20 year old over 10 years at 2% interest rate =", epvt)
```

## Pension Valuation

<details>
  <summary>What is a pension and pension valuation?</summary>
  <p style="font-weight: bold; color: blue;">
  Pension valuation is the process of determining the present value of future pension benefits that employees or retirees are entitled to receive from their employers or pension funds. It involves estimating the current worth of these future payments, taking into account factors such as inflation, mortality rates, interest rates, and the expected duration of payments.
   </p>
</details>

<details>
  <summary>Purpose and Factors Considered?</summary>
  <p style="font-weight: bold; color: blue;">
 Purpose: The primary goal of pension valuation is to assess the financial obligations of a pension plan or fund accurately. This helps pension managers, actuaries, and policymakers understand the funding requirements necessary to meet future pension obligations.

Factors Considered: Key factors considered in pension valuation include:

Discount Rates: Used to discount future cash flows to their present value, reflecting the time value of money and investment returns.

Mortality Assumptions: Expected longevity of pension recipients affects the duration and total amount of pension payments.

Inflation Rates: Adjustments for inflation ensure that future pension payments maintain their purchasing power.

Plan Design: Specific features of the pension plan, such as retirement age, payment options, and eligibility criteria, influence valuation calculations.
   </p>
</details>



**Calculating the PV of a pension at age 65**

The pension pays 20,000 a year from age 65 to age 100 (36 payments at the start of each year), increasing by 2% a year for inflation. Interest is 4% a year after 65 and 3% a year between 20 and 65, a simple non-flat interest-rate assumption. The PV ignores mortality; the EPV then accounts for the chance of dying before each payment.

```{r}
benefits <- 20000 * 1.02 ^ (0:35)




discount_factors <- 1.04 ^ - (0:35)




PV_65 <- sum(benefits * discount_factors)
cat("PV of pension at age 65 =",PV_65)



PV_20 <- PV_65 * 1.03 ^ - 45
cat("PV of pension at age 20 =",PV_20)
```


**EPV of pension at age 20**

```{r}
# Survival probabilities of (65) up to age 100
px <- 1 - life_table_2021$qx
kpx <- c(1, cumprod(px[(65 + 1):(99 + 1)]))




EPV_65 <- sum(benefits * discount_factors * kpx)
cat("PV and EPV of pension at age 65 =", PV_65, EPV_65, "\n")




EPV_20 <- EPV_65 * (1.03 ^ - 45 * prod(px[(20 + 1):(64 + 1)]))
cat("PV and EPV of pension at age 20 =", PV_20, EPV_20)
```

### Retirement Plan

You're planning for your retirement at age 40. You want to buy an annuity that will provide you 10000 annually for life, starting at age 65. If you die before the annuity starts, you get paid nothing. Interest is 3% a year. The single premium is the EPV of the benefits.

```{r}
# Survival probabilities of (40)

kpx <- c(1, cumprod(px[(40 + 1):length(px)]))



discount_factors <- (1 + 0.03) ^ -(0:(length(kpx) - 1))




benefits <- c(rep(0, 25), rep(10000, length(kpx) - 25))




single_premium <- sum(benefits * discount_factors * kpx)
cat("Single premium =",single_premium)

```

Suppose you want to finance this deferred life annuity with annual premiums payable for 25 years beginning at age 40 (only while you are alive). However, you plan to reduce your work hours from age 55 onwards, therefore the premium will reduce by 1 half after 15 years. What is the initial premium you will pay

```{r}
# Premium pattern rho
rho <- c(rep(1, 15), rep(0.5, 10), rep(0, length(kpx) - 25))


initial_premium <- single_premium / sum(rho * discount_factors * kpx, na.rm = TRUE)
cat("Initial premium =", initial_premium, "\n")


annual_premiums <- initial_premium * rho


filtered_annual_premiums <- annual_premiums[annual_premiums != 0]
cat("Filtered annual premiums =", filtered_annual_premiums, "\n")

# Sum of the annual premiums (no actuarial discounting)
sap <- sum(filtered_annual_premiums)
cat("Sum of annual premiums =", sap, "\n")

```

Find out if this product would be a favorable deal for you. What is the present value of what you would receive if you die at 75, or at 95, compared with the single premium of `r round(single_premium, 2)`?

```{r}
# Curtate life expectancy of (40)
cat("Curtate life expectancy of (40):", sum(kpx[-1]), "years\n")

# PV of benefits received if (40) dies at a given age (payments at 65, 66, ..., up to that age)
pv_if_die_at <- function(death_age) {
  k <- 0:(death_age - 40)
  sum(benefits[k + 1] * discount_factors[k + 1])
}
cat("PV of benefits if death at 75 =", pv_if_die_at(75), "\n")
cat("PV of benefits if death at 95 =", pv_if_die_at(95), "\n")

# Break-even: first age at death where the PV of benefits exceeds the single premium
breakeven_age <- (65:109)[which(sapply(65:109, pv_if_die_at) >= single_premium)[1]]
cat("Break-even age at death:", breakeven_age)
```

If you die before about `r breakeven_age`, you get back less (in present value) than the `r round(single_premium, 2)` you paid; if you live longer, you get back more. Dying at 75 returns `r round(pv_if_die_at(75))`, a loss, and living to 95 returns `r round(pv_if_die_at(95))`, a gain. The premium is the expected value of the benefits, so averaged over everyone who buys it the product is a fair deal (before any insurer profit or expenses). Whether it is a good deal for *you* depends on whether you expect to live longer than average.


#### Integrating Generalized Linear Models (GLMs) and Time Series Analysis

**Using GLMs to Model Mortality: the Gompertz Law**

Adult mortality follows a well-known pattern called the Gompertz law: the death rate grows exponentially with age, so log(mx) is close to a straight line in age. That makes it a natural GLM: a Poisson model for the deaths dx, with a log link, Age as the predictor and log(Lx) (person-years lived) as an offset, so that the fitted quantity is the death rate mx = dx / Lx.

We fit ages 30-100 of the 2021 table. Below 30, mortality has an infant and young-adult accident "hump" that a straight line can't capture. Above 100, the HMD values are themselves smoothed by a model rather than observed directly.

One caveat applies to every model in this section: a life table's dx and Lx are scaled to a hypothetical 100,000 births, not real counts. The fitted rates are valid, but the standard errors, p-values and AIC values describe that hypothetical population, so read them as relative rather than absolute. Real counts are in the HMD `Deaths_1x1` and `Exposures_1x1` files.

```{r}
adults_2021 <- subset(life_table_2021, Age >= 30 & Age <= 100)
gompertz <- glm(dx ~ Age + offset(log(Lx)), family = poisson, data = adults_2021)
coef(gompertz)

b <- coef(gompertz)[["Age"]]
cat("Mortality rises by", round(100 * (exp(b) - 1), 1), "% per year of age;",
    "it doubles every", round(log(2) / b, 1), "years\n")

plot(adults_2021$Age, adults_2021$mx, log = "y", pch = 20, col = "blue",
     xlab = "Age", ylab = expression(paste("Death rate ", m[x], " (log scale)")),
     main = "Actual vs Gompertz GLM mortality rates (Canada, 2021)")
lines(adults_2021$Age, fitted(gompertz) / adults_2021$Lx, col = "red", lwd = 2)
legend("topleft", legend = c("Actual", "Gompertz GLM"), col = c("blue", "red"),
       pch = c(20, NA), lty = c(NA, 1), lwd = c(NA, 2))
```

From about age 55 upward the fitted line tracks the data closely: mortality roughly doubles every `r round(log(2) / b, 1)` years of age, the classic Gompertz result. Between 30 and 55 the actual rates sit above the line, because deaths at those ages include more accidents, overdoses and other causes that don't rise exponentially with age. The Poisson fit is driven by where most deaths happen (older ages), so it follows those ages most closely.

*Why the earlier version of this chart looked wrong:* it plotted all 101 years (1921-2021) in one `lines()` call. Each time the line reached age 110 for one year, it jumped back to age 0 for the next year, and those jumps were the diagonal bands. It also fitted qx as a logistic curve that is straight in age, which can't represent mortality that rises exponentially with age and also has an infant spike. That's why the red curve missed badly.

**Adding calendar year: mortality improvement**

Adding Year to the model measures how fast mortality has fallen over time at a given age.

```{r}
adults_all <- subset(life_table_all, Age >= 30 & Age <= 100)
gompertz_year <- glm(dx ~ Age + Year + offset(log(Lx)), family = poisson, data = adults_all)
coef(gompertz_year)
improvement <- 1 - exp(coef(gompertz_year)[["Year"]])
cat("Average mortality improvement:", round(100 * improvement, 2), "% per year\n")

show_years <- c(1921, 1971, 2021)
cols <- c("darkorange", "darkgreen", "blue")
plot(NULL, xlim = c(30, 100), ylim = range(adults_all$mx), log = "y",
     xlab = "Age", ylab = expression(paste("Death rate ", m[x], " (log scale)")),
     main = "Actual (points) vs GLM (lines) by year")
for (j in seq_along(show_years)) {
  yr <- subset(adults_all, Year == show_years[j])
  points(yr$Age, yr$mx, pch = 20, cex = 0.6, col = cols[j])
  lines(yr$Age, predict(gompertz_year, newdata = yr, type = "response") / yr$Lx, col = cols[j], lwd = 2)
}
legend("topleft", legend = show_years, col = cols, lwd = 2, title = "Year")
```

On average, adult mortality fell by about `r round(100 * improvement, 1)`% a year. The model assumes the same improvement at every age and in every year, which is too simple, and the chart shows where it breaks down. In 1921 the actual rates sit above the line at younger ages and below it at older ages, meaning younger adults have since improved faster than average and the oldest more slowly. In 2021 the actual rates sit above the line at most ages, because improvement has slowed in recent years and 2021 was a pandemic year. The GAM below relaxes these assumptions.

**A flexible model: a GAM over age and year**

A generalized additive model (GAM) replaces the straight lines with a smooth surface over age and year, so improvement can differ by age. It uses the same Poisson setup, now over ages 0-100. An earlier version also used mx as a predictor; that is circular, because qx is calculated from mx, so the model was just recovering its own input.

```{r}
library(mgcv)
ages_0_100 <- subset(life_table_all, Age <= 100)
# Smooth on log(age + 1): mortality changes very fast in the first few years of life,
# and the log scale stretches those ages out so the smooth can follow them
ages_0_100$log_age <- log(ages_0_100$Age + 1)
gam_model <- gam(dx ~ te(log_age, Year, k = c(20, 8)) + offset(log(Lx)), family = poisson, data = ages_0_100)
cat("Deviance explained:", round(100 * summary(gam_model)$dev.expl, 2), "%\n")

plot(NULL, xlim = c(0, 100), ylim = range(ages_0_100$mx[ages_0_100$mx > 0]), log = "y",
     xlab = "Age", ylab = expression(paste("Death rate ", m[x], " (log scale)")),
     main = "Actual (points) vs GAM (lines) by year")
for (j in seq_along(show_years)) {
  yr <- subset(ages_0_100, Year == show_years[j])
  points(yr$Age, yr$mx, pch = 20, cex = 0.6, col = cols[j])
  lines(yr$Age, fitted(gam_model)[ages_0_100$Year == show_years[j]] / yr$Lx, col = cols[j], lwd = 2)
}
legend("bottomright", legend = show_years, col = cols, lwd = 2, title = "Year")
```

The GAM follows the whole age pattern, including infant mortality and the young-adult hump, and lets each age improve at its own pace. The largest gaps are at ages 5-15, where death rates are so low (around 1 in 10,000) that the yearly values are noisy.

# Life Insurance

You have bought a life insurance that is sold to 20 year olds and will pay 10000 at the end of the year of death if you die at age 30 (between your 30th and 31st birthdays). The payment would be made at age 31, 11 years from now. Using an interest rate of 1%, how much is the expected present value (EPV)?

```{r}
qx <- life_table_2021$qx
px <- 1 - qx

# 10-year survival probability of (20)
kpx <- prod(px[(20 + 1):(29 + 1)])
cat("10p20 =", kpx, "\n")

# 10-year deferred mortality probability of (20)
kqx <- kpx * qx[30 + 1]
cat("10|q20 =", kqx, "\n")

# Discount factor
discount_factor <- (1 + 0.01) ^ - 11
cat("v^11 =", discount_factor, "\n")

# EPV of the simple life insurance
cat("EPV of simple life insurance =",10000 * discount_factor * kqx)
```

## Whole, Temporary and Deferred Life Insurance

### Function to find EPV of a whole life Insurance

```{r}

whole_life_insurance <- function(age, i, lt) {
  q <- lt$qx[(age + 1):nrow(lt)]              # q_x, ..., q_110 (= 1)
  kpx <- c(1, cumprod(1 - q))[seq_along(q)]   # k p_x for k = 0, 1, ...
  kqx <- kpx * q                              # k|q_x: dies in year k + 1
  sum((1 + i) ^ -seq_along(q) * kqx)          # paid at the end of the year of death
}

# Function to plot EPV by age
plot_by_age <- function(lt, interest_rate = 0.03) {
  ages <- 0:(nrow(lt) - 1)
  epvs <- sapply(ages, function(age) whole_life_insurance(age, interest_rate, lt))
  plot(ages, epvs, type = "l", col = "blue", 
       main = "EPV of Whole Life Insurance by Age", 
       xlab = "Age", ylab = "EPV", lwd = 2)
}

# Function to plot EPV by interest rate for a given age
plot_by_interest_rate <- function(lt, target_age = 20) {
  interest_rates <- seq(0.01, 0.1, by = 0.01)
  epvs <- sapply(interest_rates, function(i) whole_life_insurance(target_age, i, lt))
  plot(interest_rates, epvs, type = "l", col = "red", 
       main = paste("EPV of Whole Life Insurance for Age", target_age, "by Interest Rate"), 
       xlab = "Interest Rate", ylab = "EPV", lwd = 2)
}


plot_by_age(life_table_2021, interest_rate = 0.03)

plot_by_interest_rate(life_table_2021, target_age = 20)

# Check against the annuity: for whole life, A_x = 1 - d * (annuity due), with d = i / (1 + i)
i_chk <- 0.03
cat("A_20 =", whole_life_insurance(20, i_chk, life_table_2021),
    "   1 - d * annuity due =", 1 - i_chk / (1 + i_chk) * life_annuity_due(20, i_chk, life_table_2021))
```

## Temporary Life Insurance

**Finding the EPV of a life insurance where the coverage is restricted in time for a 20 year old in Canada in 2021 using an interest rate of 2%**

```{r}
# Function to compute the EPV of a temporary life insurance
temporary_life_insurance <- function(age, n, i, lt) {
  q <- lt$qx[(age + 1):nrow(lt)]
  n <- min(n, length(q))
  kpx <- c(1, cumprod(1 - q))[1:n]
  kqx <- kpx * q[1:n]                         # deaths in the first n years only
  sum((1 + i) ^ -(1:n) * kqx)
}

cat("EPV of a temporary life insurance for (20) over a period of 45 years at interest rate 2% =",temporary_life_insurance(20, 45, 0.02, life_table_2021))

cat(" EPV of a whole life insurance for (20) at interest rate 2% =",whole_life_insurance(20, 0.02, life_table_2021))
```

## Deferred Life Insurance

**Calculating EPV of a deferred life insurance for constant interest rate on age x**

<details>
  <summary><span style="color:blue;">Explanation of Calculating EPV of a deferred life insurance for constant interest rate on age x</span></summary>
  In this code, we are calculating the Expected Present Value (EPV) of a deferred whole life insurance policy. The EPV is calculated by considering the mortality rates (`qx`), survival probabilities (`kpx`), and discount factors based on a constant interest rate (`i`). The deferred period (`u`) is taken into account to determine when the insurance benefits start to be paid. The EPV is the sum of the discounted expected benefits.
</details>

```{r}
# Function to compute the EPV of a deferred whole life insurance
deferred_life_insurance <- function(age, u, i, lt) {
  q <- lt$qx[(age + 1):nrow(lt)]
  kpx <- c(1, cumprod(1 - q))[seq_along(q)]
  kqx <- kpx * q
  covered <- seq_along(q) > u                 # no cover for deaths in the first u years
  sum(covered * (1 + i) ^ -seq_along(q) * kqx)
}


cat("EPV of a deferred life insurance for (20) deferred over 45 years at interest rate 2% =",deferred_life_insurance(20, 45, 0.02, life_table_2021), "\n")

# Check: 45-year temporary cover + cover deferred 45 years = whole life cover
cat("Temporary + deferred =", temporary_life_insurance(20, 45, 0.02, life_table_2021) + deferred_life_insurance(20, 45, 0.02, life_table_2021),
    "   Whole life =", whole_life_insurance(20, 0.02, life_table_2021))
```

**Suppose you are age 48 and want to insure a benefit of 40000 for death between the ages of 55 and 75 (paid at the end of the year of death). Assume interest rate is 5%**

<details>
  <summary><span style="color:blue;">Explanation of insuring a benefit of 40000 for death between the ages of 55 and 75</span></summary>
  In this code, we are calculating the Expected Present Value (EPV) of death benefits for an insurance policy. The policyholder is currently 48 years old and will receive a benefit of 40000 if they die between the ages of 55 and 75. We use mortality rates (`qx`), survival probabilities (`kpx`), and discount factors based on an interest rate (`i`) to calculate the EPV of these benefits.
</details>

```{r}
# Deferred mortality probabilites of (48)
kqx <- c(1, cumprod(px[(48 + 1):(73 + 1)])) * qx[(48 + 1):(74 + 1)]
i <- 0.05


discount_factors <- (1 + i) ^ - (1:length(kqx))


benefits <- c(rep(0, 7), rep(40000, length(kqx) - 7))


EPV_death_benefits <- sum(benefits * discount_factors * kqx)
cat("EPV of death benefits =",EPV_death_benefits)
```

**Now suppose you want to add a savings component to the policy that will add 80000 if you are alive at age 75. You want to finance this endowment insurance with constant premiums.**

<details>
  <summary><span style="color:blue;">Explanation of adding a savings component to the policy with constant premiums</span></summary>
  This code calculates the Expected Present Value (EPV) of adding a savings component to an insurance policy. The policyholder will receive an additional 80000 if they are alive at age 75. We calculate the EPV of the pure endowment and then the level annual premium, paid at the start of each year from 48 to 74 while alive, that finances both. That premium is the EPV of the benefits divided by the EPV of a premium of 1 a year.
</details>

```{r}
# Pure endowment
EPV_pure_endowment <- 80000 * (1 + i) ^ - 27 * prod(px[(48 + 1):(74 + 1)])
cat(" EPV of pure endowment =",EPV_pure_endowment)

# Premium pattern
kpx <- c(1, cumprod(px[(48 + 1):(73 + 1)]))
discount_factors <- (1 + i) ^ - (0:(length(kpx) - 1))
rho <- rep(1, length(kpx))
EPV_rho <- sum(rho * discount_factors * kpx)
cat("Premium pattern EPV = ",EPV_rho)

# Premium level
cat("Premium level =", (EPV_death_benefits + EPV_pure_endowment) / EPV_rho)
```

## Shiny App for Life Insurance Calculator

<details>
  <summary><span style="color:blue;">Explanation of Shiny App for Life Insurance Calculator</span></summary>
  This Shiny app allows users to calculate the Expected Present Value (EPV) of different types of life insurance policies (whole life, temporary, or deferred). The user inputs their age, interest rate, and the type of insurance, and the app calculates the EPV using predefined functions for each insurance type.
</details>

```{r, eval = FALSE}
library(shiny)

ui <- fluidPage(
  titlePanel("Life Insurance Calculator"),
  sidebarLayout(
    sidebarPanel(
      numericInput("age", "Age:", 20, min = 0, max = 110),
      numericInput("interest", "Interest Rate:", 0.02, min = 0, max = 0.2, step = 0.001),
      selectInput("type", "Insurance Type:", 
                  choices = c("Whole Life" = "whole", 
                              "Temporary" = "temp", 
                              "Deferred" = "deferred")),
      conditionalPanel(
        condition = "input.type == 'temp' || input.type == 'deferred'",
        numericInput("term", "Term/Deferment (years):", 10, min = 1, max = 50)
      )
    ),
    mainPanel(
      h3("Expected Present Value:"),
      textOutput("epv")
    )
  )
)

server <- function(input, output) {
  output$epv <- renderText({
    epv <- switch(input$type,
                  "whole" = whole_life_insurance(input$age, input$interest, life_table_2021),
                  "temp" = temporary_life_insurance(input$age, input$term, input$interest, life_table_2021),
                  "deferred" = deferred_life_insurance(input$age, input$term, input$interest, life_table_2021))
    paste0("$", round(epv, 2))
  })
}

shinyApp(ui = ui, server = server)


```

*The live Shiny app can't run on this static GitHub Pages site. Open `index.Rmd` in RStudio, run the earlier chunks, then run this chunk to launch it. Sample results:*

```{r, echo = FALSE}
grid <- expand.grid(Age = c(20, 40, 65), Interest = c(0.02, 0.05))
grid$`Whole life` <- mapply(whole_life_insurance, grid$Age, grid$Interest, MoreArgs = list(lt = life_table_2021))
grid$`Temporary (20 yrs)` <- mapply(temporary_life_insurance, grid$Age, 20, grid$Interest, MoreArgs = list(lt = life_table_2021))
grid$`Deferred (20 yrs)` <- mapply(deferred_life_insurance, grid$Age, 20, grid$Interest, MoreArgs = list(lt = life_table_2021))
knitr::kable(grid, digits = 4, caption = "EPV of a death benefit of 1 (2021 Canada table)")
```



### Integrating Survival Models

#### Kaplan-Meier Estimator for Survival Probabilities

<details>
  <summary><span style="color:blue;">Explanation of Kaplan-Meier Estimator for Survival Probabilities</span></summary>
  The Kaplan-Meier estimator works on individual lifetimes. A life table row is not a person: it says that dx of the 100,000 people died at age x. So we expand the 2021 table into deaths at each age, weighted by dx, and fit Kaplan-Meier to that. Because no one is censored, the estimate should reproduce the life table's own survival curve, lx / 100,000, which makes a good check.
</details>

```{r}
library(survival)

# Each age is one group of deaths, weighted by how many of the 100,000 died at that age
deaths_2021 <- data.frame(age_at_death = life_table_2021$Age + 1,  # died before the next birthday
                          weight = life_table_2021$dx)
deaths_2021 <- subset(deaths_2021, weight > 0)
km_fit <- survfit(Surv(age_at_death) ~ 1, data = deaths_2021, weights = weight)

plot(km_fit, conf.int = FALSE, xlab = "Age", ylab = "Survival Probability",
     main = "Kaplan-Meier Estimate of Survival Probability (Canada, 2021)")
points(life_table_2021$Age, life_table_2021$lx / 1e5, pch = 20, cex = 0.5, col = "red")
legend("bottomleft", legend = c("Kaplan-Meier", "Life table lx / 100,000"),
       col = c("black", "red"), lty = c(1, NA), pch = c(NA, 20))


```


#### Cox Proportional Hazards Model

We can use the Cox Proportional Hazards model to analyze the effect of covariates on survival time.

<details>
  <summary><span style="color:blue;">Explanation of Cox Proportional Hazards Model</span></summary>
  A Cox model compares the hazard (the instantaneous risk of death) between groups. Life tables have no individual-level covariates, so we compare two calendar periods instead: the 1951 and 2021 tables, each expanded into weighted deaths the same way as for Kaplan-Meier. The coefficient gives the hazard ratio, how many times higher or lower the risk of death is at the same age in 2021 than in 1951. An earlier version used Age both as the survival time and as a covariate, and treated each table row as a person, so its output had no meaning.
</details>

```{r}
deaths_by_year <- function(yr) {
  t <- subset(life_table_all, Year == yr & dx > 0)
  data.frame(age_at_death = t$Age + 1, weight = t$dx, period = factor(yr))
}
cox_data <- rbind(deaths_by_year(1951), deaths_by_year(2021))
cox_data$period <- relevel(cox_data$period, ref = "1951")

cox_model <- coxph(Surv(age_at_death) ~ period, data = cox_data, weights = weight)
hr <- exp(coef(cox_model))
cat("Hazard ratio, 2021 vs 1951:", round(hr, 3), "\n")

km_two <- survfit(Surv(age_at_death) ~ period, data = cox_data, weights = weight)
plot(km_two, col = c("darkorange", "blue"), lwd = 2, xlab = "Age", ylab = "Survival Probability",
     main = "Survival curves: 1951 vs 2021 (Canada, both sexes)")
legend("bottomleft", legend = c("1951", "2021"), col = c("darkorange", "blue"), lwd = 2)

# The proportional-hazards assumption: compare the actual mortality ratio by age
ratio_by_age <- subset(life_table_all, Year == 2021 & Age <= 100)$mx /
                subset(life_table_all, Year == 1951 & Age <= 100)$mx
plot(0:100, ratio_by_age, type = "l", log = "y", xlab = "Age",
     ylab = "m(x) in 2021 / m(x) in 1951 (log scale)", main = "Is the hazard ratio constant by age?")
abline(h = hr, col = "red", lty = 2)
legend("bottomright", legend = c("Actual ratio", "Cox hazard ratio"), col = c("black", "red"), lty = c(1, 2))
```

The Cox model says that, averaged over ages, the risk of death in 2021 was about `r round(100 * (1 - hr))`% lower than in 1951. The second chart shows why this should be read with care: the actual ratio isn't constant. Mortality fell by about 90% in childhood, by roughly 35-60% at adult ages, and by only about 25% at age 100, so the proportional-hazards assumption doesn't hold well here. A single hazard ratio is a summary, not a rule for every age.



## Integrating Credibility Theory

#### Simple Credibility Model

**We will use a simple credibility model to blend new mortality experience with a prior expectation.**

Here the new (observed) experience is 2021, which was affected by the COVID-19 pandemic, and the prior (expected) experience is the 2015-2019 average. The credibility-weighted rate is Z × observed + (1 − Z) × expected, where Z, the credibility factor, says how much weight the new data gets.

<details>
  <summary><span style="color:blue;">Explanation of Simple Credibility Model</span></summary>
  This code demonstrates a simple credibility model to adjust mortality rates based on observed and expected data. The credibility model combines new experience (the observed 2021 rates) with prior experience (the expected 2015-2019 rates) to produce adjusted rates. The credibility factor Z determines the weight given to the observed data.
</details>

```{r}
# Observed experience: the 2021 table. Expected (prior) experience: the
# average 2015-2019 table, i.e. pre-pandemic mortality.
filtered_life_table <- life_table_all %>%
  filter(Year >= 2015, Year <= 2019) %>%
  group_by(Age) %>%
  summarise(qx = mean(qx), .groups = "drop") %>%
  as.data.frame()

# Ensure observed_qx and expected_qx cover the same age range
common_ages <- intersect(life_table_2021$Age, filtered_life_table$Age)

# Filter both life_table_2021 and filtered_life_table to only include common ages
life_table_2021_filtered <- life_table_2021[life_table_2021$Age %in% common_ages, ]
filtered_life_table_filtered <- filtered_life_table[filtered_life_table$Age %in% common_ages, ]

# Extract qx values for common ages
observed_qx <- life_table_2021_filtered$qx
expected_qx <- filtered_life_table_filtered$qx

cat("Summary of Ages in life_table_2021_filtered:\n")
print(summary(life_table_2021_filtered$Age))

cat("Summary of Ages in filtered_life_table_filtered:\n")
print(summary(filtered_life_table_filtered$Age))


if (length(observed_qx) != length(expected_qx)) {
  stop("Lengths of observed_qx and expected_qx do not match after alignment!")
}

# Apply credibility model
simple_credibility <- function(observed_qx, expected_qx, credibility_factor) {
  adjusted_qx <- credibility_factor * observed_qx + (1 - credibility_factor) * expected_qx
  return(adjusted_qx)
}

credibility_factor <- 0.8
adjusted_qx <- simple_credibility(observed_qx, expected_qx, credibility_factor)
cat("Ages where 2021 mortality was above the 2015-2019 average:", sum(observed_qx > expected_qx), "of", length(observed_qx), "\n")


# Plot the adjusted mortality rates
plot(life_table_2021_filtered$Age, adjusted_qx, type = "l", col = "green", xlab = "Age", ylab = "Mortality Rate (log scale)", main = "Adjusted Mortality Rates using Credibility Theory", log = "y")
lines(life_table_2021_filtered$Age, observed_qx, col = "blue")
lines(life_table_2021_filtered$Age, expected_qx, col = "grey50", lty = 2)
legend("topleft", legend = c("Adjusted", "Observed (2021)", "Expected (2015-2019 avg)"), col = c("green", "blue", "grey50"), lty = c(1, 1, 2))
```

**Choosing Z.** Z = 0.8 is a judgment input here, used to show the mechanics, and the sample results table further down shows how the answer changes with Z. In practice Z comes from how much data stands behind the observed rates. For example, the limited-fluctuation rule gives full credibility (Z = 1) at about 1,082 deaths, and otherwise Z = √(deaths / 1,082). That requires real death counts, which this life table (scaled to 100,000 births) doesn't contain. On the log scale the three curves look close. At about half of all ages, the 2021 rate is within 5% of the pre-pandemic average, but at some adult ages it is up to `r round(100 * (max(observed_qx[31:91] / expected_qx[31:91]) - 1))`% higher.



### Incorporating Adjusted Mortality Rates into Life Insurance Calculations

**We will now use the adjusted mortality rates in the life insurance calculations.**

<details>
  <summary><span style="color:blue;">Explanation of using adjusted mortality rates in life insurance calculations</span></summary>
  This code uses the adjusted mortality rates obtained from the credibility model to calculate the Expected Present Value (EPV) of a whole life insurance policy. The EPV is calculated by considering the adjusted mortality rates (`adjusted_qx`), survival probabilities (`kpx`), and discount factors based on a constant interest rate (`i`). The adjusted EPV lies between the EPVs from the 2021 rates and from the 2015-2019 rates, closer to 2021 because Z = 0.8.
</details>

```{r}
# Function to compute the EPV of a whole life insurance with adjusted mortality rates
whole_life_insurance_adjusted <- function(age, i, adjusted_qx) {
  # same calculation as whole_life_insurance(), using the adjusted rates
  whole_life_insurance(age, i, data.frame(qx = adjusted_qx))
}

# EPV of whole life insurance for a 20-year-old at 2% interest rate using adjusted mortality rates
epv_adjusted <- whole_life_insurance_adjusted(20, 0.02, adjusted_qx)
cat("EPV of whole life insurance with adjusted mortality rates =", epv_adjusted, "\n")
cat("Using 2021 rates only =", whole_life_insurance(20, 0.02, life_table_2021),
    "  using 2015-2019 rates only =", whole_life_insurance(20, 0.02, data.frame(qx = expected_qx)))


```


## Shiny App for Life Insurance Calculator with Adjusted Mortality Rates

```{r, eval = FALSE}
library(shiny)

ui <- fluidPage(
  titlePanel("Life Insurance Calculator with Adjusted Mortality Rates"),
  sidebarLayout(
    sidebarPanel(
      numericInput("age", "Age:", 20, min = 0, max = 110),
      numericInput("interest", "Interest Rate:", 0.02, min = 0, max = 0.2, step = 0.001),
      numericInput("credibility", "Credibility Factor:", 0.8, min = 0, max = 1, step = 0.1),
      selectInput("type", "Insurance Type:", 
                  choices = c("Whole Life" = "whole", 
                              "Temporary" = "temp", 
                              "Deferred" = "deferred")),
      conditionalPanel(
        condition = "input.type == 'temp' || input.type == 'deferred'",
        numericInput("term", "Term/Deferment (years):", 10, min = 1, max = 50)
      )
    ),
    mainPanel(
      h3("Expected Present Value:"),
      textOutput("epv")
    )
  )
)

server <- function(input, output) {
  output$epv <- renderText({
    adjusted_qx <- simple_credibility(observed_qx, expected_qx, input$credibility)
    adjusted_table <- data.frame(Age = life_table_2021_filtered$Age, qx = adjusted_qx)
    epv <- switch(input$type,
                  "whole" = whole_life_insurance_adjusted(input$age, input$interest, adjusted_qx),
                  "temp" = temporary_life_insurance(input$age, input$term, input$interest, adjusted_table),
                  "deferred" = deferred_life_insurance(input$age, input$term, input$interest, adjusted_table))
    paste0("$", round(epv, 2))
  })
}

shinyApp(ui = ui, server = server)


```

*The live Shiny app can't run on this static GitHub Pages site. Open `index.Rmd` in RStudio, run the earlier chunks, then run this chunk to launch it. Sample results:*

```{r, echo = FALSE}
cred <- c(0, 0.5, 0.8, 1)
res <- data.frame(`Credibility factor` = cred, check.names = FALSE)
for (a in c(20, 40, 65)) {
  res[[paste0("Whole life, age ", a)]] <- sapply(cred, function(z)
    whole_life_insurance_adjusted(a, 0.02, simple_credibility(observed_qx, expected_qx, z)))
}
knitr::kable(res, digits = 4, caption = "EPV of whole life insurance of 1 at 2% using credibility-adjusted rates")
```









## Generalized Linear Models (GLMs) for Mortality Rates

**We will use GLMs with various link functions to model mortality rates.**

<details>
  <summary style="font-weight: bold; color: blue;">Generalized Linear Models (GLMs) for Mortality Rates</summary>
  <p>
    <strong>Overview:</strong><br>
    We model qx as a binomial probability (dx deaths out of lx alive) with Age and Year as predictors, and compare three link functions, which are three different shapes for how the predictors map to a probability. We use ages 30-100 for the same reasons as the Gompertz model above. An earlier version also used mx as a predictor; qx is calculated directly from mx, so that just fed the answer into the model.<br><br>
    <strong>1. Logit:</strong> log(q / (1 − q)) is linear in the predictors. This is ordinary logistic regression.<br><br>
    <strong>2. Probit:</strong> the predictors shift a standard normal quantile.<br><br>
    <strong>3. Complementary log-log:</strong> log(−log(1 − q)) is linear. Since −log(1 − q) is the cumulative hazard over the year, this is equivalent to a log-linear (Gompertz-type) model for the force of mortality, so it is the natural choice for mortality.<br><br>
    <strong>Model comparison:</strong> lower AIC means a better fit. Because lx is scaled to 100,000 births, the AIC values only make sense relative to each other, not on their own.
</details>

```{r}
adults_all <- subset(life_table_all, Age >= 30 & Age <= 100)

glm_logit   <- glm(cbind(dx, lx - dx) ~ Age + Year, family = binomial(link = "logit"),   data = adults_all)
glm_probit  <- glm(cbind(dx, lx - dx) ~ Age + Year, family = binomial(link = "probit"),  data = adults_all)
glm_cloglog <- glm(cbind(dx, lx - dx) ~ Age + Year, family = binomial(link = "cloglog"), data = adults_all)

# Compare the models using AIC
aics <- AIC(glm_logit, glm_probit, glm_cloglog)
aics
best_glm <- list(glm_logit, glm_probit, glm_cloglog)[[which.min(aics$AIC)]]
cat("Best model by AIC:", rownames(aics)[which.min(aics$AIC)], "\n")
summary(best_glm)$coefficients

# Actual vs predicted, drawn separately for each year so lines don't join across years
plot(NULL, xlim = c(30, 100), ylim = range(adults_all$qx), log = "y",
     xlab = "Age", ylab = "Probability of Death (log scale)",
     main = "Actual (points) vs Predicted (lines) Mortality Rates (GLM)")
for (j in seq_along(show_years)) {
  yr <- subset(adults_all, Year == show_years[j])
  points(yr$Age, yr$qx, pch = 20, cex = 0.6, col = cols[j])
  lines(yr$Age, predict(best_glm, newdata = yr, type = "response"), col = cols[j], lwd = 2)
}
legend("topleft", legend = show_years, col = cols, lwd = 2, title = "Year")


```

The complementary log-log link fits best, as expected, since it corresponds to mortality growing exponentially with age.


## ARIMA/ARMA Models for Mortality Rate Forecasting

**We will use an ARIMA model to forecast life expectancy at birth.**

An earlier version averaged qx over all ages 0-110 each year and forecast that. That average isn't a meaningful mortality measure: it gives the same weight to age 109 as to age 30, so it is dominated by the very old ages where qx is 0.3 to 1. Life expectancy at birth summarises the whole year's mortality in one number that people understand, so we forecast that instead.

<details>
  <summary><span style="color:blue;">Explanation of the Code and Graph</span></summary>
  <p>1. <strong>Series:</strong> life expectancy at birth (e0) for each year 1921-2021, stored as an annual time series.</p>
  <p>2. <strong>Model:</strong> <code>auto.arima()</code> chooses the ARIMA orders by AIC. Life expectancy trends upward rather than fluctuating around a fixed level, so the model works with differences of the series (the "I" in ARIMA) instead of the raw values.</p>
  <p>3. <strong>Forecast:</strong> 10 years ahead, with 80% and 95% prediction intervals shown as shaded bands.</p>
</details>

```{r}
library(forecast)

e0_series <- subset(life_table_all, Age == 0)
e0_ts <- ts(e0_series$ex, start = min(e0_series$Year))
arima_model <- auto.arima(e0_ts)
summary(arima_model)

forecasts_arima <- forecast(arima_model, h = 10)
plot(forecasts_arima, main = "Forecast of Life Expectancy at Birth (ARIMA)", xlab = "Year",
     ylab = expression(e[0]), fcol = "red")
lines(e0_ts, col = "blue")
legend("bottomright", legend = c("Historical", "Forecast"), col = c("blue", "red"), lty = 1)
forecasts_arima
```

**How we interpret these results**:

The blue line is life expectancy at birth from 1921 to 2021, and the red line is the forecast for the next 10 years, with 80% and 95% prediction intervals shaded. `auto.arima()` chose ARIMA(`r paste(arimaorder(arima_model), collapse = ",")`), which continues the recent upward trend from the 2021 starting point. That starting point is depressed by the pandemic, and a pure time-series model can't know whether that dip is temporary. This is one reason actuaries use structured models such as Lee-Carter for mortality projections.

<!-- ## Bühlmann and Bühlmann-Straub Models for Credibility Theory -->
<!-- Work in progress, not rendered. Note before re-enabling: in buhlmann(), var(m) is the variance of a
     single number (NA). Bühlmann needs several risks, each observed several times: the expected process
     variance is the average within-risk variance, and the variance of the hypothetical means is the
     variance between risk means minus EPV / n. actuar::cm() implements both models. -->


<!-- # ```{r} -->
<!-- # # Load the life tables -->
<!-- # life_table <- read.table("C:\\Users\\megacrazyleo\\Desktop\\SQL\\R\\bltper_1x1.txt", header = FALSE, skip = 3, sep = "") -->
<!-- # colnames(life_table) <- c("Year", "Age", "mx", "qx", "ax", "lx", "dx", "Lx", "Tx", "ex") -->
<!-- #  -->
<!-- # # Subset the life table for the year 2021 -->
<!-- # life_table_2021 <- subset(life_table, Year == 2021) -->
<!-- #  -->
<!-- # # Ensure Age is numeric and remove NAs -->
<!-- # life_table$Age <- as.numeric(life_table$Age) -->
<!-- # life_table_2021$Age <- as.numeric(life_table_2021$Age) -->
<!-- #  -->
<!-- # # Remove rows with NA ages -->
<!-- # life_table <- life_table[!is.na(life_table$Age), ] -->
<!-- # life_table_2021 <- life_table_2021[!is.na(life_table_2021$Age), ] -->
<!-- #  -->
<!-- # # Print summary of ages to diagnose potential issues -->
<!-- # cat("Summary of Ages in life_table:\n") -->
<!-- # print(summary(life_table$Age)) -->
<!-- #  -->
<!-- # cat("Summary of Ages in life_table_2021:\n") -->
<!-- # print(summary(life_table_2021$Age)) -->
<!-- #  -->
<!-- # # Filter the historical life table to only include data for 2021 ages -->
<!-- # common_ages <- intersect(life_table$Age, life_table_2021$Age) -->
<!-- # life_table_filtered <- life_table[life_table$Age %in% common_ages & life_table$Year != 2021, ] -->
<!-- # life_table_2021_filtered <- life_table_2021[life_table_2021$Age %in% common_ages, ] -->
<!-- #  -->
<!-- # # Sort by Age to ensure alignment -->
<!-- # life_table_filtered <- life_table_filtered[order(life_table_filtered$Age), ] -->
<!-- # life_table_2021_filtered <- life_table_2021_filtered[order(life_table_2021_filtered$Age), ] -->
<!-- #  -->
<!-- # # Extract qx values for the common ages -->
<!-- # expected_qx <- life_table_filtered$qx -->
<!-- # observed_qx <- life_table_2021_filtered$qx -->
<!-- #  -->
<!-- # # Print lengths and summaries for diagnostic purposes -->
<!-- # cat("Length of observed_qx:", length(observed_qx), "\n") -->
<!-- # cat("Length of expected_qx:", length(expected_qx), "\n") -->
<!-- #  -->
<!-- # cat("First few values of observed_qx:\n") -->
<!-- # print(head(observed_qx)) -->
<!-- #  -->
<!-- # cat("First few values of expected_qx:\n") -->
<!-- # print(head(expected_qx)) -->
<!-- #  -->
<!-- # # Ensure they have the same length -->
<!-- # if (length(observed_qx) != length(expected_qx)) { -->
<!-- #   stop("Lengths of observed_qx and expected_qx do not match after alignment!") -->
<!-- # } -->
<!-- #  -->
<!-- # # Define the Bühlmann and Bühlmann-Straub models -->
<!-- # buhlmann <- function(x) { -->
<!-- #   n <- length(x) -->
<!-- #   m <- mean(x) -->
<!-- #   v <- var(x) -->
<!-- #   z <- n / (n + v / var(m)) -->
<!-- #   return(z * x + (1 - z) * m) -->
<!-- # } -->
<!-- #  -->
<!-- # buhlmann_straub <- function(x, w) { -->
<!-- #   n <- length(x) -->
<!-- #   m <- weighted.mean(x, w) -->
<!-- #   v <- sum(w * (x - m)^2) / sum(w) -->
<!-- #   z <- sum(w) / (sum(w) + v / var(m)) -->
<!-- #   return(z * x + (1 - z) * m) -->
<!-- # } -->
<!-- #  -->
<!-- # # Apply Bühlmann Model -->
<!-- # adjusted_qx_buhlmann <- buhlmann(observed_qx) -->
<!-- #  -->
<!-- # # Apply Bühlmann-Straub Model -->
<!-- # weights <- rep(1, length(observed_qx)) # Assuming equal weights for simplicity -->
<!-- # adjusted_qx_buhlmann_straub <- buhlmann_straub(observed_qx, weights) -->
<!-- #  -->
<!-- # # Plot adjusted mortality rates -->
<!-- # plot(life_table_2021_filtered$Age, observed_qx, type = "l", col = "blue", xlab = "Age", ylab = "Mortality Rate", main = "Adjusted Mortality Rates using Credibility Theory") -->
<!-- # lines(life_table_2021_filtered$Age, adjusted_qx_buhlmann, col = "green") -->
<!-- # lines(life_table_2021_filtered$Age, adjusted_qx_buhlmann_straub, col = "red") -->
<!-- # legend("topright", legend = c("Observed", "Bühlmann", "Bühlmann-Straub"), col = c("blue", "green", "red"), lty = 1) -->
<!-- #  -->
<!-- #  -->
<!-- #  -->
<!-- #  -->
<!-- #  -->
<!-- # ``` -->

<!-- ### Incorporating Adjusted Mortality Rates into Life Insurance Calculations -->

<!-- # ```{r} -->
<!-- # # Function to compute the EPV of a whole life insurance with adjusted mortality rates -->
<!-- # whole_life_insurance_adjusted <- function(age, i, adjusted_qx) { -->
<!-- #   px <- 1 - adjusted_qx -->
<!-- #   kpx <- c(1, cumprod(px[(age + 1):(length(px) - 1)])) -->
<!-- #   kqx <- kpx * adjusted_qx[(age + 1):length(adjusted_qx)] -->
<!-- #   discount_factors <- (1 + i) ^ - (1:length(kqx)) -->
<!-- #   sum(discount_factors * kqx, na.rm = TRUE) -->
<!-- # } -->
<!-- # -->
<!-- # # EPV of whole life insurance for a 20-year-old at 2% interest rate using adjusted mortality rates -->
<!-- # epv_adjusted_buhlmann <- whole_life_insurance_adjusted(20, 0.02, adjusted_qx_buhlmann) -->
<!-- # cat("EPV of whole life insurance with Bühlmann adjusted mortality rates =", epv_adjusted_buhlmann) -->
<!-- # -->
<!-- # epv_adjusted_buhlmann_straub <- whole_life_insurance_adjusted(20, 0.02, adjusted_qx_buhlmann_straub) -->
<!-- # cat("EPV of whole life insurance with Bühlmann-Straub adjusted mortality rates =", epv_adjusted_buhlmann_straub) -->
<!-- # -->
<!-- # ``` -->






</div>



